---
title: "MBCO under multiple imputation (D4-stacked)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MBCO under multiple imputation (D4-stacked)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_deps <- requireNamespace("mice", quietly = TRUE) &&
  requireNamespace("medfit", quietly = TRUE) &&
  requireNamespace("RMediation", quietly = TRUE)
knitr::opts_chunk$set(eval = has_deps)
```

## Why MBCO does not commute with Rubin's rules

The **model-based constrained optimization** (MBCO) test of mediation tests
\(H_0: a b = 0\). Because \(a b = 0\) iff \(a = 0\) **or** \(b = 0\), the
constrained log-likelihood is a **branch union** --- the better of the
"drop the a-path" and "drop the b-path" fits:

\[
T = 2\left[\ell_{\text{full}} - \max(\ell_{a=0},\, \ell_{b=0})\right].
\]

That `max()` is non-linear, so the MBCO statistic of the *pooled* estimate is
**not** the pool of the per-imputation MBCO statistics. You cannot pool first and
test second. Instead, missingmed keeps every per-imputation fit (exposed by
`per_imputation_list()`) and combines the **likelihood-ratio statistics** with
the **D4** rule (Chan & Meng, 2022; Grund, Lüdtke & Robitzsch, 2021):

\[
d_S = \frac{\text{LRT(stacked data)}}{K}, \quad
r_4 = \max\!\left(0, \tfrac{K+1}{k(K-1)}(\bar d - d_S)\right), \quad
D_4 = \frac{d_S}{k(1 + r_4)} \sim F_{k,\nu}.
\]

## In practice

```{r demo}
library(missingmed)

set.seed(2026)
n <- 300
C <- rnorm(n)
X <- rbinom(n, 1, plogis(0.3 * C))
M <- 0.39 * X + 0.3 * C + rnorm(n)
Y <- 0.2 * X + 0.0 * M + 0.3 * C + rnorm(n) # interior null: b = 0
d <- data.frame(X = X, M = M, Y = Y, C = C)
d$M[sample(n, 45)] <- NA
d$Y[sample(n, 30)] <- NA

imp <- mice::mice(d, m = 20, method = "norm", printFlag = FALSE)
fit <- run(set_md_mediation(imp, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M"))

infer(fit, type = "mbco", ariv = "fixed")
```

## Which branch feeds \(r_4\): the `ariv` argument

Each imputation's statistic has its own winning branch, and the imputations need
not agree. `ariv` decides which statistics enter \(\bar d\), and so \(r_4\):

- `ariv = "fixed"` (the default) recomputes every imputation's statistic on the
  branch the **stacked** constrained fit chose. All imputations then test the
  same constraint, with the same \(k\).
- `ariv = "own"` uses each imputation's own winning branch, the standard Chan
  & Meng \(r_4\). It reproduces missingmed 0.4.0 on full-rank designs, and it
  refuses to pool when the winning branches remove different numbers of
  parameters (for example, with an `X:M` term in the outcome model).

The result records the branch diagnostics:

```{r ariv}
r_fixed <- infer(fit, type = "mbco", ariv = "fixed")
r_fixed@stacked_branch # "a" (a = 0) or "b" (b = 0)
r_fixed@branch_mix # do the imputations disagree on their own branch?
r_fixed@p_branch_a # share of imputations whose own branch is a = 0
tidy(r_fixed)
```

When no imputation disagrees with the stacked fit, the two agree. Reporting the
diagnostics needs both single-path null fits in every imputation, which is why
each imputation is fit three times.

The same test runs on any list of completed data frames, without the pipeline:

```{r mbco-d4}
implist <- mice::complete(imp, "all")
mbco_d4(implist, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", ariv = "fixed"
)
```

The per-imputation fits MBCO needs are available directly:

```{r accessor}
acc <- per_imputation_list(fit)
acc$m
length(acc$per_imputation)
```

Asking for MBCO on the *pooled* result is an error by design:

```{r error, error = TRUE}
infer(pool(fit), type = "mbco")
```

## References

- Chan, K. W., & Meng, X.-L. (2022). Multiple improvements of multiple
  imputation likelihood ratio tests. *Statistica Sinica*.
- Grund, S., Lüdtke, O., & Robitzsch, A. (2021). Pooling methods for
  likelihood-ratio tests with multiply imputed data. *Psychological Methods*.
