---
title: "Testing an indirect effect with incomplete data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing an indirect effect with incomplete data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_deps <- requireNamespace("mice", quietly = TRUE) &&
  requireNamespace("medfit", quietly = TRUE) &&
  requireNamespace("RMediation", quietly = TRUE)
knitr::opts_chunk$set(eval = has_deps)
```

## What you get

One fitted object from `run()` gives two complementary results for the
indirect effect \(a b\):

- `infer(fit, type = "mc")`: a pooled **Monte Carlo confidence interval** for
  \(a b\) (`"mc"` is the default `type`).
- `infer(fit, type = "mbco", ariv = "fixed")`: the **D4-MBCO likelihood-ratio
  test** of \(H_0: a b = 0\) under multiple imputation.

Report the interval when the question is how large the indirect effect is.
Report the test when the claim is that there is an indirect effect at all.
Every MBCO call in this vignette names `ariv` explicitly; section
[Options and edge cases](#options-and-edge-cases) says what it does.

## Before you impute: the imputation model

missingmed does not impute. It takes a `mice::mids` object, and the test and
the interval are only as good as the imputation model behind it. Five points
matter most for mediation:

- **Congeniality.** The imputation model must include every variable in the
  analysis: the treatment, the mediator, the **outcome** and every covariate.
  An imputation model that leaves \(Y\) out imputes \(M\) as if it were
  unrelated to \(Y\), which biases \(b\) toward 0.
- **Interactions.** If the outcome model has an `X:M` term, the imputation
  model must carry that product too. There are two ways to do it in `mice`:
  - *Passive imputation*: add a column `XM` and give it the method
    `"~I(X * M)"`, so it is recomputed from the imputed `X` and `M` at every
    iteration and always equals their product. The mediator's own imputation
    model should then not use `XM` as a predictor (it is a function of \(M\)).
  - *Impute the product as its own variable*: give `XM` an ordinary method.
    The imputed product is then free to differ from `X * M`. Note that
    missingmed's formula `Y ~ X * M + C` recomputes the product from the
    imputed `X` and `M`.

  See the *Passive imputation* section of `?mice::mice` and van Buuren (2018)
  for both. The [interaction example](#models-with-an-xm-interaction) below
  uses passive imputation.
- **Auxiliary variables.** Include variables that predict missingness, or that
  predict the incomplete variables, even when they are not in the analysis
  model.
- **Number of imputations.** Use \(K \ge 20\), and more when the fraction of
  missing information is high. D4 needs at least two imputations;
  `infer(type = "mbco")` refuses \(K = 1\).
- **Reproducibility.** Pass `seed =` to `mice::mice()`.

## Worked example

The data are synthetic: a binary treatment `X`, a mediator `M`, an outcome
`Y` and a fully observed covariate `C`, with \(a = 0.5\), \(b = 0.3\) and
\(n = 400\). `M` and `Y` go missing at random, with probabilities that depend
on `C`. Because the missingness depends on an observed variable, the complete
cases are not a random subsample, and imputation using `C` is needed.

```{r data}
library(missingmed)

set.seed(8644)
n <- 400
C <- rnorm(n)
X <- rbinom(n, 1, 0.5)
M <- 0.5 * X + 0.3 * C + rnorm(n)
Y <- 0.2 * X + 0.3 * M + 0.3 * C + rnorm(n)
d <- data.frame(X = X, M = M, Y = Y, C = C)

# MAR: higher C makes M more likely to be missing, lower C makes Y more likely
d$M[runif(n) < plogis(-1.8 + C)] <- NA
d$Y[runif(n) < plogis(-1.8 - C)] <- NA
round(colMeans(is.na(d)), 2)
```

Impute with all four variables in the model, then fit the mediation model in
every imputation:

```{r impute}
imp <- mice::mice(d, m = 20, method = "norm", seed = 3051, printFlag = FALSE)

md <- set_md_mediation(imp, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M"
)
fit <- run(md)
```

## The interval

`infer(type = "mc")` pools the per-imputation estimates with Rubin's rules
and draws a Monte Carlo interval for \(a b\) from the pooled estimates.
Called on the fit, it pools first; `infer(pool(fit), type = "mc")` gives the
same interval up to Monte Carlo error.

```{r mc}
infer(fit, type = "mc", level = 0.95, n.mc = 1e5)
```

`Estimate` is the mean of the Monte Carlo draws of \(a b\) (it differs from
\(\hat a \hat b\) by the pooled covariance of \(\hat a\) and \(\hat b\)),
and `CI` is the 95% interval.
`MC.Error` is the Monte Carlo error of the estimate; raise `n.mc` if it
matters at the precision you report.

## The test

```{r mbco}
res <- infer(fit, type = "mbco", ariv = "fixed")
res
```

How to read it:

- `D4` is the pooled test statistic, referred to an \(F(k, \nu)\)
  distribution. \(k\) is the number of parameters the null constraint
  removes: 1 in this model, for either branch.
- `p` is the p-value of \(H_0: a b = 0\).
- `r4` is the relative increase in variance due to the missing data, the
  between-imputation inflation that deflates `D4`. It is floored at 0.
- `nu` is the denominator degrees of freedom, \(k(K-1)(1 + 1/r_4)^2\) (Chan &
  Meng, 2022, eq. 2.15 of arXiv:1711.08822, the same as `mitml::testModels(method = "D4")`). It is
  `Inf` when `r4` is 0.
- `d_S` is the likelihood-ratio statistic of the stacked data, divided by
  \(K\).

### The result object

The result is an S7 `MbcoMIResult`. It is still a named numeric vector
`c(D4, p, r4, nu, d_S)`, so the usual indexing works, and the diagnostics are
properties:

```{r result}
res["p"]
res[["p"]]
res@stacked_branch # the null the stacked constrained fit chose
res@branch_mix # do the imputations' own winning branches differ?
res@p_branch_a # share of imputations whose own winning branch is a = 0
tidy(res)
```

`res$p` is not valid: use `res["p"]` or `res[["p"]]`.

The null \(a b = 0\) holds when \(a = 0\) **or** \(b = 0\), so the
constrained fit is the better of two single-path fits, one per branch.
`stacked_branch` says which branch won on the stacked data. Each imputation
also has its own winning branch; `p_branch_a` is the share of imputations
whose own winner is \(a = 0\), and `branch_mix` is `TRUE` when the imputations
do not all agree.

## Options and edge cases

### `ariv = "fixed"` or `"own"`

`ariv` sets which per-imputation statistics enter \(r_4\):

- `ariv = "fixed"`, the default since missingmed 0.5.0,
  recomputes every imputation's statistic on the branch the **stacked**
  constrained fit chose. All imputations then test the same constraint, with
  the same \(k\).
- `ariv = "own"` lets each imputation use its own winning branch. On full-rank
  designs its `D4` and `r4` equal those of missingmed 0.4.0 and earlier (`nu` and
  `p` differ slightly, because the denominator degrees of freedom now follow
  Chan & Meng). **It can be liberal**: see *Operating characteristics*. Prefer
  `"fixed"` unless you need to reproduce an earlier analysis.

The two give the same result when every imputation's own branch matches the
stacked fit's. Reporting `branch_mix` and `p_branch_a` needs both single-path
null fits in every imputation, so each imputation is fit three times under
either option.

### Without the pipeline: `mbco_d4()`

The same test runs on any list of completed data frames:

```{r mbco-d4}
implist <- mice::complete(imp, "all")
mbco_d4(implist, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", ariv = "fixed"
)
```

For a structural equation model, give a lavaan model string and the outcome name
instead of the two formulas (maximum likelihood only; `model` cannot be combined
with the formula or family arguments):

```{r mbco-d4-model}
mbco_d4(implist,
  model = "M ~ X + C\nY ~ M + X + C",
  treatment = "X", mediator = "M", outcome = "Y"
)
```

### A single imputation

D4 needs at least two imputations, so \(K = 1\) is an informative error. For
a single complete dataset, use complete-data MBCO in RMediation (Tofighi &
Kelley, 2020).

```{r k1, error = TRUE}
mbco_d4(implist[1], Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", ariv = "fixed"
)
```

### When \(k\) differs across imputations

Under `ariv = "fixed"` every imputation is tested with the stacked fit's
\(k\), the number of parameters the chosen branch's constraint removes. When
the constraint removes a different number in some imputation than in the
stacked data, for example when a level of a factor that interacts with the
treatment or mediator is absent from that imputation, `infer()` stops and
names the imputation. The remedy is to drop or merge the sparse level.

### MBCO on a pooled object

MBCO needs the per-imputation fits, so asking for it on the *pooled* result
is an error by design (see [below](#why-mbco-does-not-commute-with-rubins-rules)):

```{r error, error = TRUE}
infer(pool(fit), type = "mbco", ariv = "fixed")
```

## Models with an X:M interaction

The outcome model may include a treatment-by-mediator interaction,
`Y ~ X * M + C`. The imputation model must then carry the product; here it is
imputed passively, and `XM` is removed as a predictor of `M`:

```{r interaction}
d2 <- d
d2$XM <- d2$X * d2$M
meth <- mice::make.method(d2)
meth[c("M", "Y")] <- "norm"
meth["XM"] <- "~I(X * M)"
pred <- mice::make.predictorMatrix(d2)
pred["M", "XM"] <- 0
imp2 <- mice::mice(d2,
  m = 20, method = meth, predictorMatrix = pred,
  seed = 3053, printFlag = FALSE
)

fit2 <- run(set_md_mediation(imp2, Y ~ X * M + C, M ~ X + C,
  treatment = "X", mediator = "M"
))
res2 <- infer(fit2, type = "mbco", ariv = "fixed")
res2
res2@k
```

For the \(b = 0\) branch, the null drops **every** term that contains `M`:
here `M` and `X:M`, so that branch removes \(k = 2\) parameters, while the
\(a = 0\) branch removes 1. Under `ariv = "fixed"` all imputations use the
stacked fit's branch, so there is a single \(k\) and the test is well
defined; `res2@k` reports it.
Under `ariv = "own"`, imputations whose own winning branches differ remove
different numbers of parameters, and the test refuses to pool.

With an X:M term the indirect effect is \(a (b + \theta_3 x)\), where
\(\theta_3\) is the X:M coefficient: the effect of a one-unit increase in
`X` through `M`, with `X` held at \(x\) in the outcome model. It depends on
\(x\), so the Monte Carlo interval needs `treatment_level`. For a 0/1
treatment, `treatment_level = 1` gives the total natural indirect effect and
`treatment_level = 0` the pure natural indirect effect:

```{r interaction-mc}
infer(fit2, type = "mc", treatment_level = 1)
infer(fit2, type = "mc", treatment_level = 0)
```

Without `treatment_level`, `infer(type = "mc")` stops and asks for it. MBCO
needs no such choice: its null, \(a = 0\) or \(b = \theta_3 = 0\), removes
the indirect effect at every \(x\).

## Why MBCO does not commute with Rubin's rules

The **model-based constrained optimization** (MBCO) test of mediation tests
\(H_0: a b = 0\). Because \(a b = 0\) iff \(a = 0\) **or** \(b = 0\), the
constrained log-likelihood is a **branch union**: the better of the
"drop the a-path" and "drop the b-path" fits,

\[
T = 2\left[\ell_{\text{full}} - \max(\ell_{a=0},\, \ell_{b=0})\right].
\]

That `max()` is non-linear, so the MBCO statistic of the *pooled* estimate is
**not** the pool of the per-imputation MBCO statistics. You cannot pool first
and test second. Instead, missingmed keeps every per-imputation fit and
combines the **likelihood-ratio statistics** with the **D4** rule (Chan &
Meng, 2022; Grund, Lüdtke & Robitzsch, 2023):

\[
d_S = \frac{\text{LRT(stacked data)}}{K}, \quad
r_4 = \max\!\left(0, \tfrac{K+1}{k(K-1)}(\bar d - d_S)\right), \quad
D_4 = \frac{d_S}{k(1 + r_4)} \sim F_{k,\nu}, \quad
\nu = k(K-1)\left(1 + \frac{1}{r_4}\right)^2,
\]

where \(\bar d = K^{-1} \sum_{i} d_{i}\) averages the per-imputation
statistics. The two `ariv` options differ only in \(d_{i}\). Let \(B\) be the
branch the stacked constrained fit chose, \(a = 0\) or \(b = 0\). Then

\[
d_{i}^{\text{fixed}} = 2\left[\ell_{\text{full}, i} - \ell_{B, i}\right],
\qquad
d_{i}^{\text{own}} = 2\left[\ell_{\text{full}, i}
  - \max(\ell_{a=0, i},\, \ell_{b=0, i})\right],
\]

with \(\ell_{\cdot, i}\) the log-likelihood in imputation \(i\). `"fixed"`
evaluates every imputation at the same constraint \(B\); `"own"` takes each
imputation's maximum.

The per-imputation fits are available directly:

```{r accessor}
acc <- per_imputation_list(fit)
acc$m
length(acc$per_imputation)
```

## Operating characteristics

The test is the smaller of two likelihood-ratio statistics, so it is **conservative**
rather than exact when the null is true on a boundary. In a simulation of the
lavaan models (40 settings, 1000 replications each, \(n\) = 200 and 500, 25% and 40%
of the mediator or its indicators missing, 20 imputations, normal data, 5% level):

| True paths | Rejection rate |
|---|---|
| one path null, the other strong (\(a = 0, b = .3\) or \(a = .3, b = 0\)) | 3.3% to 5.7%, mean 4.2% |
| one path null, the other weak (.1) | 0.7% to 2.2% |
| both null | 0.1% to 0.4% |

No setting rejected more than 6.5% of the time. The cost of the conservatism is
power near \(a = b = 0\): a small indirect effect is rejected less often than a
test that was exact there would reject it.

The glm engine has its own grid (132 settings, 1000 replications each, 20
imputations unless noted, MAR missingness in the mediator, 5% level, default
`ariv = "fixed"`): Gaussian models including an `X:M` term (\(k = 2\) on one
branch), binary and Poisson outcomes, a binary mediator, a rare binary outcome
with \(n\) = 150 and 300, \(n\) = 200 and 500, 25% and 40% missing, and 5 and 10
imputations at \(n\) = 200.

| True paths | Rejection rate (`"fixed"`) |
|---|---|
| one path null, the other strong | 1.4% to 5.7%, mean 3.5% |
| one path null, the other weak (.1) | 0.1% to 2.4%, mean 1.0% |
| both null | 0.1% to 0.7%, mean 0.3% |

No setting rejected more than 5.7% of the time, including 5 and 10 imputations.
Power at \(a = b = .3\) (\(n\) = 500) ranged from 43% (binary mediator) to 100%
(Gaussian with `X:M`). Refits that did not converge were refused in 14 of 132,000
replications, all in the rare-outcome settings.

**`ariv = "own"` is less well behaved.** On the same data it rejected a true null
more than 6.5% of the time in 23 of the 80 non-Gaussian null settings, up to 13.8%
(Poisson outcome, \(n\) = 200, 40% missing), and in a separate grid of 20
plain-Gaussian settings (\(k = 1\) on both branches) up to 8.3% (\(n\) = 200, 40%
missing; `"fixed"`: at most 4.7%). The excess grows with the share of missing data
and shrinks with \(n\). The reported Gaussian-with-`X:M` figure for `"own"` is not
comparable: there `"own"` errors whenever the imputations' winning branches remove
different numbers of parameters, which happened in half of the replications.

This grid tested the MBCO test only. Coverage of the Monte-Carlo interval for a
model with an `X:M` term (`type = "mc"` with `treatment_level`) was not
calibrated.

## Not covered here

- Models with a latent variable or a lavaan model string: the same test runs with
  `engine = "lavaan"` (maximum likelihood only); see
  `vignette("lavaan-sem")` (the *Structural equation models* article).
- FIML-based MBCO.
- IPW: `method = "ipw"` has no MBCO test.

## References

- Chan, K. W., & Meng, X.-L. (2022). Multiple improvements of multiple
  imputation likelihood ratio tests. *Statistica Sinica*.
  <https://doi.org/10.5705/ss.202019.0314>
- Grund, S., Lüdtke, O., & Robitzsch, A. (2023). Pooling methods for
  likelihood ratio tests in multiply imputed data sets. *Psychological
  Methods*, *28*(5), 1207--1221. <https://doi.org/10.1037/met0000556>
- Tofighi, D., & Kelley, K. (2020). Improved inference in mediation analysis:
  Introducing the model-based constrained optimization procedure.
  *Psychological Methods*, *25*(4), 496--515.
  <https://doi.org/10.1037/met0000259>
- van Buuren, S. (2018). *Flexible imputation of missing data* (2nd ed.).
  Chapman and Hall/CRC. <https://doi.org/10.1201/9780429492259>
