---
title: "Getting started with missingmed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with missingmed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# These deps are not on CRAN; only evaluate when available.
has_deps <- requireNamespace("mice", quietly = TRUE) &&
  requireNamespace("medfit", quietly = TRUE) &&
  requireNamespace("RMediation", quietly = TRUE)
knitr::opts_chunk$set(eval = has_deps)
```

**missingmed** runs SEM-based mediation across multiply imputed datasets and
pools with Rubin's rules. It is a thin orchestration layer: it **fits** each
imputation with [medfit](https://data-wise.github.io/medfit/) and delegates
**inference** to [RMediation](https://data-wise.github.io/rmediation/). The S7
pipeline is four verbs:

```
set_md_mediation()  ->  run()  ->  pool()  ->  infer()
   MDMediationData    MDMediationFit  MDMediationResult   CI / MBCO
```

## A worked example

We simulate a simple mediation model `X -> M -> Y` with a confounder `C`, impose
some MAR missingness, and impute with `mice`.

```{r setup}
library(missingmed)

set.seed(2026)
n <- 300
C <- rnorm(n)
X <- rbinom(n, 1, plogis(0.3 * C))
M <- 0.5 * X + 0.3 * C + rnorm(n)
Y <- 0.2 * X + 0.4 * M + 0.3 * C + rnorm(n)
d <- data.frame(X = X, M = M, Y = Y, C = C)
d$M[sample(n, 45)] <- NA
d$Y[sample(n, 30)] <- NA

imp <- mice::mice(d, m = 20, method = "norm", printFlag = FALSE)
```

### 1. Specify the mediation model

`set_md_mediation()` records the outcome and mediator formulas plus the
treatment/mediator roles.

```{r set}
md <- set_md_mediation(
  imp,
  formula_y = Y ~ X + M + C,
  formula_m = M ~ X + C,
  treatment = "X",
  mediator  = "M"
)
md
```

### 2. Fit each imputation

`run()` fits every imputed dataset with `medfit::fit_mediation()`, yielding a
list of **named** `medfit::MediationData` objects.

```{r run}
fit <- run(md)
fit
```

### 3. Pool with Rubin's rules

`pool()` combines the per-imputation estimates and variance-covariance into a
single **named** pooled `medfit::MediationData`.

```{r pool}
res <- pool(fit)
summary(res)
```

### 4. Inference on the indirect effect

The Monte-Carlo / distribution-of-the-product CI is computed by RMediation on
the pooled object:

```{r mc}
infer(res, type = "mc")
```

For the **MBCO** likelihood-ratio test of `H0: a*b = 0`, use the fit object (see
`vignette("mbco-mi")` for why pooling does not commute with MBCO):

```{r mbco}
infer(fit, type = "mbco")
```

## A binary mediator (or outcome)

`set_md_mediation()` takes `family_m` and `family_y` and passes them to
`medfit::fit_mediation()`, whose engine is a GLM. Nothing else in the pipeline
changes -- the same four verbs handle a binary mediator:

```{r glm}
set.seed(7)
Mb <- rbinom(n, 1, plogis(-0.2 + 1.2 * X + 0.3 * C))
Yb <- 0.2 * X + 1.0 * Mb + 0.3 * C + rnorm(n)
db <- data.frame(X = X, M = Mb, Y = Yb, C = C)
db$M[sample(n, 45)] <- NA

impb <- mice::mice(db, m = 10, printFlag = FALSE)
mdb <- set_md_mediation(impb, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M",
  family_m = binomial()
)
infer(pool(run(mdb)), type = "mc")
```

**Mind the scale.** With a non-identity link, `a` and `b` are link-scale
coefficients, so `a * b` and its interval are on the link scale as well -- here,
a change in `Y` per log-odds of `M`. It is not a risk difference or an odds
ratio, and exponentiating it does not make one. `vignette("technical")` has the
full table and the reasoning.

## Migrating from the S4 API

The S4 entry points (`set_sem()`, `run_sem()`, `pool_sem()`) are **deprecated**
in favor of the S7 verbs above. They still work for one release cycle but emit
a deprecation warning.
