---
title: "Technical reference: design, contracts, and methodology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical reference: design, contracts, and methodology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette is the single technical reference for **missingmed**: the
architecture, the cross-package contracts it depends on, the statistical
methodology of every estimator and inference path, and the engineering decisions
(and their rationale) made along the way. Code blocks are illustrative
(`eval = FALSE`); runnable walkthroughs live in `vignette("missingmed")` and
`vignette("mbco-mi")`.

---

# 1. Architecture (S7, four verbs, three classes)

missingmed is a **thin orchestration layer**: it runs mediation across the
"missing-data middle" and delegates the statistics outward — **fitting** to
[medfit](https://data-wise.github.io/medfit/), **inference** to
[RMediation](https://data-wise.github.io/rmediation/), and (later) simulation to
medsim. It is written in **S7** to match the newer ecosystem packages and the
shared `medfit::MediationData` contract.

The pipeline is four verbs over three classes:

```
set_md_mediation()  ->  run()          ->  pool()             ->  infer()
   MDMediationData       MDMediationFit     MDMediationResult      CI / MBCO
   (data + spec)         (per-imputation    (pooled named          (RMediation
                          named Med.Data)    Med.Data)              or hosted D4)
```

The same pipeline as a diagram:

<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 880 225" width="100%" role="img" aria-label="missingmed pipeline: set_md_mediation to run to pool to infer">
<title>missingmed pipeline</title>
<defs><marker id="ah" viewBox="0 0 10 10" refX="9" refY="5" markerWidth="7" markerHeight="7" orient="auto-start-reverse"><path d="M0,0 L10,5 L0,10 z" fill="#444"/></marker></defs>
<g font-family="-apple-system,Segoe UI,Helvetica,Arial,sans-serif">
<rect x="15" y="30" width="160" height="52" rx="7" fill="#eef4fb" stroke="#4a76a8" stroke-width="1.4"/>
<text x="95.0" y="51" text-anchor="middle" font-size="12.5" font-weight="600" fill="#1b2b3a">MDMediationData</text>
<text x="95.0" y="68" text-anchor="middle" font-size="10.5" fill="#5a6b7a">data + spec</text>
<rect x="245" y="30" width="160" height="52" rx="7" fill="#eef4fb" stroke="#4a76a8" stroke-width="1.4"/>
<text x="325.0" y="51" text-anchor="middle" font-size="12.5" font-weight="600" fill="#1b2b3a">MDMediationFit</text>
<text x="325.0" y="68" text-anchor="middle" font-size="10.5" fill="#5a6b7a">m named MediationData</text>
<rect x="475" y="30" width="160" height="52" rx="7" fill="#eef4fb" stroke="#4a76a8" stroke-width="1.4"/>
<text x="555.0" y="51" text-anchor="middle" font-size="12.5" font-weight="600" fill="#1b2b3a">MDMediationResult</text>
<text x="555.0" y="68" text-anchor="middle" font-size="10.5" fill="#5a6b7a">pooled MediationData</text>
<rect x="705" y="30" width="160" height="52" rx="7" fill="#eef4fb" stroke="#4a76a8" stroke-width="1.4"/>
<text x="785.0" y="51" text-anchor="middle" font-size="12.5" font-weight="600" fill="#1b2b3a">RMediation</text>
<text x="785.0" y="68" text-anchor="middle" font-size="10.5" fill="#5a6b7a">ci_mediation_data()</text>
<rect x="245" y="150" width="160" height="52" rx="7" fill="#fdf3e7" stroke="#c98a3c" stroke-width="1.4"/>
<text x="325.0" y="171" text-anchor="middle" font-size="12.5" font-weight="600" fill="#1b2b3a">hosted D4</text>
<text x="325.0" y="188" text-anchor="middle" font-size="10.5" fill="#5a6b7a">R/mbco_mi.R</text>
<path d="M175,56 L240,56" stroke="#444" stroke-width="1.4" fill="none" marker-end="url(#ah)"/>
<text x="207" y="50" text-anchor="middle" font-size="11" fill="#333" font-style="italic">run()</text>
<path d="M405,56 L470,56" stroke="#444" stroke-width="1.4" fill="none" marker-end="url(#ah)"/>
<text x="437" y="50" text-anchor="middle" font-size="11" fill="#333" font-style="italic">pool()</text>
<path d="M635,56 L700,56" stroke="#444" stroke-width="1.4" fill="none" marker-end="url(#ah)"/>
<text x="667" y="50" text-anchor="middle" font-size="11" fill="#333" font-style="italic">infer(&quot;mc&quot;)</text>
<path d="M325,82 L325,150" stroke="#444" stroke-width="1.4" fill="none" marker-end="url(#ah)"/>
<text x="335" y="120" font-size="11" fill="#333" font-style="italic">infer("mbco")</text>
<text x="15" y="205" font-size="10.5" fill="#5a6b7a">Note: infer("mc") consumes the pooled result; infer("mbco") reaches back to the per-imputation fit (section 4.1).</text>
</g></svg>

Note the asymmetry the linear sketch above cannot show: `infer(type = "mc")`
consumes the **pooled** result, while `infer(type = "mbco")` reaches back to the
**per-imputation** fit. That is section 4.1's non-commutativity, drawn.

<!-- Mermaid source for the diagram above. The SVG is pre-rendered so the
     vignette needs no JavaScript, no CDN and no build-time browser:
flowchart LR
  D["data<br/>mids or data.frame"] --> A
  A["set_md_mediation()<br/>MDMediationData"] -->|"method = mi"| R1["run()<br/>fit each imputation"]
  A -->|"method = ipw"| R2["run()<br/>reweight complete cases"]
  R1 --> F["MDMediationFit<br/>list of m named MediationData"]
  R2 --> F
  F -->|"pool()"| P["MDMediationResult<br/>pooled named MediationData"]
  P -->|"infer(type = mc)"| C1["RMediation::ci_mediation_data()"]
  F -->|"infer(type = mbco)"| C2["hosted D4 stacking<br/>R/mbco_mi.R"]
-->

<!-- LaTeX equivalent of the diagram above, for a PDF build:
\begin{tikzpicture}[node distance=2.2cm, every node/.style={font=\small},
  box/.style={draw, rounded corners, align=center, inner sep=4pt}]
  \node[box] (A) {\texttt{set\_md\_mediation()}\\\textbf{MDMediationData}};
  \node[box, right=of A] (F) {\texttt{run()}\\\textbf{MDMediationFit}};
  \node[box, right=of F] (P) {\texttt{pool()}\\\textbf{MDMediationResult}};
  \node[box, right=of P] (C) {\texttt{infer(type="mc")}\\RMediation};
  \node[box, below=1.4cm of P] (M) {\texttt{infer(type="mbco")}\\hosted D4};
  \draw[->] (A) -- (F); \draw[->] (F) -- (P); \draw[->] (P) -- (C);
  \draw[->] (F) -- (M) node[midway, sloped, above, font=\tiny]{per-imputation};
\end{tikzpicture}
-->

| S7 class | Carries | S4 ancestor |
|---|---|---|
| `MDMediationData` | data (`mids` or `data.frame`) + mediation spec + estimator/mechanism axes | `SemImputedData` |
| `MDMediationFit` | **list** of per-imputation named `medfit::MediationData` (+ IPW `weights`) | `SemResults` |
| `MDMediationResult` | **pooled** named `medfit::MediationData` + within/between/total vcov | `PooledSEMResults` |

**Orthogonal axes.** The estimator (`method = "mi" | "ipw"`) and the model
(formulas + `engine`) are independent. The same `run() -> pool() -> infer()`
chain serves both estimators; only `run()` branches internally.

```{r}
md <- set_md_mediation(data, formula_y = Y ~ X + M + C, formula_m = M ~ X + C,
  treatment = "X", mediator = "M", method = "mi")   # or method = "ipw"
res <- pool(run(md))
infer(res, type = "mc")
```

---

# 2. Ecosystem contracts

missingmed produces and consumes objects defined elsewhere. These contracts are
load-bearing — getting the **names** right is what makes inference work.

## 2.1 `medfit::MediationData` (the fitting contract)

`fit_mediation()` / `extract_mediation()` return a `MediationData` whose
`@estimates` is a **named** numeric vector and `@vcov` a **named** matrix:

```
estimates: m_(Intercept), m_X, m_C, y_(Intercept), y_X, y_M, y_C, a, b, c_prime
```

The convenience aliases `a` (= `m_X`), `b` (= `y_M`), `c_prime` (= `y_X`) and the
matching `@vcov` dimnames are the interface every downstream consumer resolves
**by name**.

## 2.2 `RMediation` (the inference contract)

> **Namespace note:** the package is `RMediation` (capital R-M), *not*
> `rmediation` — the source repo is `data-wise/rmediation` but
> `DESCRIPTION: Package: RMediation`. Code uses `RMediation::`.

- `ci_mediation_data(mu, level, type, n.mc)` — Monte-Carlo / distribution-of-the-
  product CI from a **single named** `MediationData`. Resolves `a`/`b` strictly
  by name (errors if unnamed). This is the `type = "mc"` path.
- `mbco(h0, h1, ...)` — likelihood-ratio MBCO, currently **OpenMx-only**; it has
  no per-imputation / MI entry point (see §4).

## 2.3 Dependency direction (no cycle)

```
missingmed  ->  medfit       (fit_mediation, MediationData)
missingmed  ->  RMediation   (ci_mediation_data, medci)
RMediation  ->  medfit       (consumes MediationData)
medsim          (Suggests, later phases)
```

missingmed imports both medfit and RMediation; neither depends back on
missingmed, so the graph is acyclic.

---

# 3. The MI estimator and Rubin's rules

Under multiple imputation, `run()` fits **every** imputed dataset with
`medfit::fit_mediation()`, yielding a list of `m` named `MediationData`. `pool()`
applies Rubin's (1987) rules on the **named** vectors so the path labels survive
pooling:

\[
\bar Q = \tfrac1m\sum_i Q_i,\quad
\bar U = \tfrac1m\sum_i U_i,\quad
B = \tfrac{1}{m-1}\sum_i (Q_i-\bar Q)(Q_i-\bar Q)^\top,\quad
T = \bar U + \left(1+\tfrac1m\right)B .
\]

### Worked example: pooling three imputations

*Time: about 3 minutes. Arithmetic only -- no calculus, no software.*

**After this you can** compute \(T\) by hand and say what fraction of your
information the missing data cost you.

**The punchline first.** Three imputations whose estimates differ by only
\(\pm 0.10\) turn a standard error of 0.100 into **0.153**. That 53% inflation
*is* the price of the missing data, and computing it is all `pool()` does.

Symbols in this example:

| symbol | plain words | here |
|---|---|---|
| \(Q_i\) | the estimate from imputation \(i\) | \(\hat a_i\) |
| \(U_i\) | how uncertain that single fit was | \(0.01\), i.e. \(SE = 0.10\) |
| \(B\) | how much the imputations disagreed | computed in step 3 |
| \(T\) | total variance, both sources | computed in step 4 |

The three fits:

| \(i\) | \(Q_i = \hat a_i\) | \(U_i\) | \(Q_i - \bar Q\) |
|---|---|---|---|
| 1 | 0.50 | 0.01 | -0.10 |
| 2 | 0.60 | 0.01 | 0.00 |
| 3 | 0.70 | 0.01 | +0.10 |

**Step 1 -- average the estimates.**
\(\bar Q = \tfrac{0.50+0.60+0.70}{3} = \mathbf{0.60}\)

**Step 2 -- average the within-imputation variances.**
\(\bar U = \tfrac{0.01+0.01+0.01}{3} = \mathbf{0.01}\)

**Step 3 -- measure the disagreement between imputations.**
\(B = \tfrac{1}{3-1}\big[(-0.10)^2 + 0^2 + (0.10)^2\big]
= \tfrac{0.02}{2} = \mathbf{0.01}\)

**Step 4 -- combine the two sources.**
\(T = \bar U + \left(1+\tfrac13\right)B = 0.01 + \tfrac43(0.01)
= \mathbf{0.0233\overline{3}}\), so \(SE = \sqrt{T} = \mathbf{0.1528}\).

Two more numbers fall straight out:

- **Fraction of missing information**
  \(\gamma = (1+\tfrac1m)B/T = 0.01\overline{3}/0.023\overline{3} =
  \mathbf{0.571}\). Over half the information about \(a\) is missing.
- **Degrees of freedom**
  \(\nu = (m-1)\big(1 + \bar U/[(1+\tfrac1m)B]\big)^2 = 2(1+0.75)^2 =
  \mathbf{6.125}\). Three imputations buy about six df, not three.

```{r rubin-decomp, eval = TRUE, echo = FALSE, fig.width = 7, fig.height = 4, fig.alt = "Stacked bar: total variance T splits into a within-imputation part 0.010 and a between-imputation part 0.0133; the standard error rises from 0.100 to 0.153."}
m <- 3; Ubar <- 0.01; B <- 0.01; corr <- (1 + 1/m) * B; T_ <- Ubar + corr
par(mar = c(4.2, 4.2, 2.6, 1), mgp = c(2.4, 0.7, 0))
b <- barplot(matrix(c(Ubar, corr), ncol = 1), col = c("#4a76a8", "#c98a3c"),
  border = NA, width = 0.35, xlim = c(0, 1.6), ylim = c(0, 0.030),
  ylab = "variance", names.arg = "T  =  Ubar  +  (1+1/m)B",
  main = sprintf("Where T comes from:  T = %.4f,  SE = %.4f", T_, sqrt(T_)))
text(b, Ubar/2, sprintf("within\n%.3f", Ubar), col = "white", cex = 0.8)
text(b, Ubar + corr/2, sprintf("between\n%.4f", corr), col = "white", cex = 0.8)
arrows(b + 0.32, 0, b + 0.32, Ubar, code = 3, length = 0.05, col = "#4a76a8")
text(b + 0.36, Ubar/2, "SE = 0.100\n(if data were complete)", pos = 4, cex = 0.75, col = "#4a76a8")
arrows(b + 0.32, 0, b + 0.32, T_, code = 3, length = 0.05, col = "#8a5a1a", lty = 2)
text(b + 0.36, T_ * 0.86, "SE = 0.153\n(+53%: the price of missingness)", pos = 4, cex = 0.75, col = "#8a5a1a")
```

**Explain this one to yourself before moving on.** Step 4 multiplies \(B\) by
\((1+\tfrac1m)\) instead of using \(B\) alone. Why should a *larger* \(m\)
make that correction *smaller*?

<details>
<summary>Answer</summary>

\(B\) is itself estimated, from only \(m\) numbers, so it carries sampling
error of its own. The \((1+\tfrac1m)\) factor inflates it to account for that.
With \(m = 3\) the factor is \(4/3\); with \(m = 100\) it is \(1.01\).
The more imputations you draw, the better you know the between-imputation
variance, and the less you must pad it.

</details>

```{r correction-vs-m, eval = TRUE, echo = FALSE, fig.width = 7, fig.height = 4, fig.alt = "The correction factor 1 + 1/m falls from 1.5 at m = 2 toward 1 as m grows, with m = 3, 10 and 20 marked."}
ms <- 2:50
par(mar = c(4.2, 4.2, 2.6, 1), mgp = c(2.4, 0.7, 0))
plot(ms, 1 + 1/ms, type = "l", lwd = 2.6, col = "#c98a3c", ylim = c(0.98, 1.55),
  xlab = "number of imputations  m", ylab = "correction factor  (1 + 1/m)",
  main = "More imputations => B better estimated => less padding")
for (k in c(3, 10, 20)) {
  points(k, 1 + 1/k, pch = 19, col = "#4a76a8", cex = 1.2)
  text(k, 1 + 1/k, sprintf("  m=%d: %.3f", k, 1 + 1/k), pos = 4, cex = 0.8)
}
abline(h = 1, lty = 3, col = "grey50")
text(42, 1.012, "limit: no padding", cex = 0.78, col = "grey30")
```

**Your turn -- same four steps, one line blanked.** Now
\(Q = (0.20,\ 0.30,\ 0.40)\), still with \(U_i = 0.01\) and \(m = 3\).
\(\bar Q = 0.30\); \(B = \underline{\hphantom{0.01}}\);
\(T = \underline{\hphantom{0.0233}}\).

<details>
<summary>Filled in</summary>

The spacing is identical to the worked case, so the deviations are again
\((-0.10, 0, +0.10)\): \(B = 0.02/2 = 0.01\) and
\(T = 0.01 + \tfrac43(0.01) = 0.0233\overline{3}\), \(SE = 0.1528\).
Only \(\bar Q\) moved. **\(T\) depends on the *spread* of the imputations,
never on their level.**

</details>

> **Why this cannot be done on \(\hat a\hat b\) directly.** Every step above
> is linear in \(Q\). The indirect effect \(ab\) is not, so `pool()`
> deliberately stops here: it pools \((a, b)\) and their covariance and hands
> the nonlinear step to `RMediation`. Pooling \(\hat a_i \hat b_i\) across
> imputations and attaching a normal interval commits two errors at once.

**You can now** read `res@tidy_table`'s `var_w`, `var_b` and `var_tot` columns
as steps 2, 3 and 4 of the arithmetic above.

The pooled `MediationData` is built by **copy-modifying** a per-imputation one
(S7 is copy-on-write), replacing only `@estimates = `\(\bar Q\) and
`@vcov = `\(T\) — so all medfit metadata (roles, predictors, families) is
inherited and the aliases stay name-addressable for `ci_mediation_data()`.

```{r}
fit <- run(md_mi)            # m named MediationData
res <- pool(fit)             # pooled named MediationData + within/between/total vcov
res@tidy_table               # term, estimate, std_error, var_w, var_b, var_tot
```

---

# 4. MBCO under MI: D4-stacking (and why it is hosted here)

## 4.1 Why MBCO does not commute with Rubin's rules

MBCO tests \(H_0: a b = 0\). Since \(a b = 0 \iff a = 0 \lor b = 0\), the
constrained log-likelihood is the **branch union**:

\[
T = 2\big[\ell_{\text{full}} - \max(\ell_{a=0},\ \ell_{b=0})\big].
\]

That `max()` is non-linear, so the MBCO statistic of the *pooled* estimate is
**not** the pool of per-imputation MBCO statistics. You cannot pool first and
test second — you need the per-imputation fits.

## 4.2 D4 combination

Hence `MDMediationFit` retains the per-imputation list (exposed by
`per_imputation_list()`), and `infer(type = "mbco")` combines the per-imputation
LRT statistics with the **D4** rule (Chan & Meng 2022; Grund et al. 2021):

\[
d_S = \tfrac{\text{LRT(stacked data)}}{K},\quad
r_4 = \max\!\Big(0, \tfrac{K+1}{k(K-1)}(\bar d - d_S)\Big),\quad
D_4 = \tfrac{d_S}{k(1+r_4)} \sim F_{k,\nu}.
\]

### Worked example: pooling three MBCO statistics

*Time: about 2 minutes.*

**After this you can** explain why averaging per-imputation test statistics is
anti-conservative, using a number.

**The punchline first.** The three statistics average to 7.5, which looks
significant. The correct pooled answer is **1.5**, which is not. The gap between
those two numbers is the missing information.

Take \(K = 3\) imputations, one constraint (\(k = 1\)):

\[
d_1 = 6.0,\quad d_2 = 7.5,\quad d_3 = 9.0
\;\Rightarrow\; \bar d = \mathbf{7.5};
\qquad d_S = \mathbf{6.0}
\]

**Step 1 -- measure the gap between the average and the stacked statistic.**
\(\bar d - d_S = 7.5 - 6.0 = 1.5\)

**Step 2 -- scale it into \(r_4\).**
\(r_4 = \max\!\big(0,\ \tfrac{K+1}{k(K-1)} \times 1.5\big)
= \tfrac42 \times 1.5 = \mathbf{3.0}\)

**Step 3 -- deflate.**
\(D_4 = \tfrac{d_S}{k(1+r_4)} = \tfrac{6.0}{1 \times 4} = \mathbf{1.5}\)

**Step 4 -- find the reference distribution.** Here \(k(K-1) = 2\), which is
not \(> 4\), so the small-sample branch applies:
\(\nu = \tfrac12 \cdot 2 \cdot (1 + \tfrac11)(1 + \tfrac13)^2 = 2(4/3)^2
= \mathbf{3.556}\), giving
\(p = \Pr[F_{1,\,3.556} > 1.5] = \mathbf{0.296}\).

```{r d4-deflation, eval = TRUE, echo = FALSE, fig.width = 7, fig.height = 4, fig.alt = "Three bars: naive mean of the per-imputation statistics 7.5, stacked statistic 6.0, and the correct D4 value 1.5."}
v <- c(7.5, 6.0, 1.5)
names(v) <- c("mean of d_k\n(naive pooling)", "stacked  d_S", "D4\n(correct)")
par(mar = c(4.2, 4.2, 2.6, 1), mgp = c(2.4, 0.7, 0))
b <- barplot(v, col = c("#b34a4a", "#c98a3c", "#4a76a8"), border = NA,
  ylim = c(0, 9), ylab = "test statistic",
  main = "Averaging per-imputation statistics is anti-conservative")
text(b, v + 0.45, sprintf("%.1f", v), cex = 1, font = 2)
segments(b[1], 7.5, b[1], 6.0, col = "grey25")
segments(b[1], 6.0, b[2], 6.0, lty = 3, col = "grey45")
text(b[1] + 0.15, 6.75, "gap 1.5 =\nmissing information\n=> r4 = 3", pos = 4, cex = 0.72, col = "grey20")
```

**Explain this one to yourself.** Why is \(\bar d\) *larger* than \(d_S\)
when both describe the same hypothesis?

<details>
<summary>Answer</summary>

Each per-imputation statistic is computed as though its filled-in dataset were
real, complete data -- so each one ignores the uncertainty that produced the
filling-in, and is optimistic. Stacking averages the log-likelihoods first, which
partly cancels that optimism. The systematic excess of \(\bar d\) over
\(d_S\) is therefore an estimate of how much of the evidence was manufactured
by the imputation model rather than observed. \(r_4\) converts that excess into
a deflation factor.

</details>

> **Watch the truncation.** If \(r_4\) hits the \(\max(0, \cdot)\) floor,
> the correction vanishes and \(D_4 = d_S/k\). Check `r4` in the returned
> object before trusting a borderline \(p\).

**You can now** say why `MDMediationFit` must retain the per-imputation list:
\(\bar d\) cannot be recovered from the pooled result alone.

## 4.3 Hosting decision

`RMediation::mbco()` is OpenMx-only with no MI entry point, so the D4 machinery
is **hosted in missingmed** (`R/mbco_mi.R`), ported from the Missing-Effect
research prototype. It reproduces the prototype **exactly** (max abs diff
\(\approx 5\times10^{-11}\) across design cells). When `RMediation` gains an MI
entry point, this code should move upstream (`TODO` noted in source).

---

# 5. The IPW estimator

IPW is a robustness complement to MI (manuscript appendix): instead of imputing,
it **reweights the observed complete cases** to represent the full sample under
MAR. The whole estimator is a single internal branch, `.ipw_run()`; `pool()` and
`infer(type = "mc")` are unchanged.

## 5.1 Pipeline

```{r}
md <- set_md_mediation(df, Y ~ X + M + C, M ~ X + C, treatment = "X",
  mediator = "M", method = "ipw",
  weight_formula = NULL, weight_stabilize = TRUE, weight_trim = 0.99,
  se_type = "sandwich")
res <- pool(run(md))         # MDMediationFit(m = 1) -> MDMediationResult (B = 0)
infer(res, type = "mc")
```

Because there is a single weighted fit, `MDMediationFit@m = 1` and `pool()`
reduces to \(\bar Q = Q_1,\ B = 0,\ T = \bar U\) — the result is structurally
identical to an MI result, so nothing downstream changes.

## 5.2 Weight estimation

Let \(R = 1\) mark a complete case over the model variables.

- **Joint complete-case model (default).** `weight_formula = NULL` fits
  \(\text{logit }P(R=1\mid Z)\) on the **fully observed** model variables (the
  MAR drivers); a single `formula` overrides the predictors.
- **Per-variable / sequential factorization.** A *named list* of formulas fits
  one model per incomplete variable and multiplies:
  \(P(\text{complete}) = \prod_V P(R_V = 1\mid \cdot)\), following the causal
  order \(X \to M \to Y\).

**Stabilized weights** (`weight_stabilize = TRUE`, default) use a treatment-only
numerator to reduce variance:
\[
w_i = \frac{\hat P(R=1\mid X_i)}{\hat P(R=1\mid Z_i)} .
\]

### Worked example: four weights, stabilized and trimmed

*Time: about 2 minutes.*

**After this you can** predict what stabilization does to a point estimate
(nothing) and what trimming does to it (moves it).

**The punchline first.** Stabilization rescales every weight by the same factor,
so it **cannot move the point estimate**. Trimming *does* move it -- and it moves
it by cutting exactly the case that needed the most correction.

Four complete cases, denominator model \(\hat P(R=1\mid Z_i)\), constant
numerator \(\hat P(R=1\mid X_i) = 0.5\):

| \(i\) | \(\hat P(R=1\mid Z_i)\) | unstabilized \(1/\hat P\) | stabilized \(0.5/\hat P\) |
|---|---|---|---|
| 1 | 0.80 | 1.250 | **0.625** |
| 2 | 0.50 | 2.000 | **1.000** |
| 3 | 0.40 | 2.500 | **1.250** |
| 4 | 0.25 | 4.000 | **2.000** |
| | mean | 2.4375 | **1.21875** |

**Read one:** case 1 had an 80% chance of being complete, so it stands in for
\(1/0.8 = 1.25\) people. Case 4 had only a 25% chance, so it stands in for
\(1/0.25 = 4\) -- it is carrying three absent people on its back.

**Two things to notice.**

1. Every stabilized weight is exactly \(0.5\times\) its unstabilized
   counterpart. A weighted regression is invariant to a common rescaling of the
   weights, so **the point estimate is identical either way**; stabilization
   buys variance, never consistency.
2. `weight_trim = 0.75` caps at the 0.75 quantile of the stabilized weights,
   \(1.25 + 0.25(2.00-1.25) = \mathbf{1.4375}\), pulling case 4 from 2.000
   down to 1.4375.

```{r ipw-weights, eval = TRUE, echo = FALSE, fig.width = 7, fig.height = 4, fig.alt = "Paired bars for four cases showing stabilized weights and the same weights after trimming; only case 4, with the lowest completeness probability, is cut."}
p4 <- c(0.80, 0.50, 0.40, 0.25)
ws <- 0.5 / p4
cap <- quantile(ws, 0.75, names = FALSE)
wt <- pmin(ws, cap)
par(mar = c(4.2, 4.2, 2.6, 1), mgp = c(2.4, 0.7, 0))
b <- barplot(rbind(ws, wt), beside = TRUE, col = c("#4a76a8", "#b34a4a"),
  border = NA, names.arg = sprintf("case %d\nP(complete)=%.2f", 1:4, p4),
  ylab = "stabilized weight", ylim = c(0, 2.4),
  main = "Trimming cuts exactly the most under-represented case")
legend("topleft", c("stabilized", "after trim (0.75 quantile)"),
  fill = c("#4a76a8", "#b34a4a"), border = NA, bty = "n", cex = 0.82)
abline(h = cap, lty = 2, col = "grey30")
text(0.8, cap + 0.12, sprintf("cap = %.4f", cap), cex = 0.78)
arrows(b[2, 4], ws[4], b[2, 4], wt[4], length = 0.06, col = "#b34a4a", lwd = 2)
```

**Explain this one to yourself.** Point 1 says a common rescaling is harmless.
Why, then, is trimming *not* harmless?

<details>
<summary>Answer</summary>

Rescaling multiplies every weight by the same constant, so the *relative*
contributions -- which is all a weighted fit sees -- are unchanged. Trimming
changes only the largest weights, so it changes the relative contributions. And
it always cuts the cases with the smallest \(\hat P(R=1\mid Z)\): exactly the
under-represented ones the weighting existed to recover. **Trimming changes the
estimand; it does not merely stabilize the estimate.**

</details>

**You can now** explain the bias-variance trade in `weight_trim` without
reaching for a simulation.

**Trimming** (`weight_trim`) caps weights at an upper quantile of the untrimmed
complete-case weights; `1` disables it.

## 5.3 Variance: sandwich vs model

IPW's weighted-GLM model-based vcov is optimistic (it ignores that weights were
estimated). `se_type = "sandwich"` (the IPW default) uses
heteroskedasticity-consistent `sandwich::vcovHC` instead. This is implemented
**in medfit** via an injectable `vcov_fun` threaded through `extract_mediation()`
(so the named-vcov assembly stays in one place); missingmed just passes
`se_type` through.

> MBCO for IPW is **not** implemented (a weighted LRT is methodologically
> distinct); `infer(type = "mbco")` on an IPW fit errors with guidance.

---

# 5A. Non-gaussian models (`family_y`, `family_m`) and the scale of `a*b`

`set_md_mediation()` takes `engine`, `family_y` and `family_m` and forwards them
to `medfit::fit_mediation()`, whose default engine already **is** `"glm"`. Both
estimators and both inference types therefore accept non-gaussian families with
no extra machinery: a binary mediator (`family_m = binomial()`), a binary
outcome (`family_y = binomial()`) and a count outcome (`family_y = poisson()`)
all run through `set_md_mediation() -> run() -> pool() -> infer()`.

## 5A.1 The scale caveat

For a non-identity link, `a` and `b` are **link-scale** coefficients, so their
product -- and the Monte-Carlo interval around it -- is on the link scale too:

| Model | `a` | `b` | `a*b` is... |
|---|---|---|---|
| gaussian / gaussian | mean difference | mean difference | a mean difference |
| binomial mediator | log-odds | mean difference | change in Y per log-odds of M |
| binomial outcome | mean difference | log-odds | log-odds units |
| poisson outcome | mean difference | log-rate | log-rate units |

Two consequences worth stating plainly:

1. `a*b` under a non-identity link is **not** a risk difference, a risk ratio or
   an odds ratio, and `exp(a*b)` does not produce one. Reporting it as such is a
   category error, not a rounding issue.
2. Rubin's rules are applied to the **coefficients**, which is standard.
   Pooling a transformed quantity (an odds ratio, say) is a different and much
   less well-behaved operation; `pool()` deliberately does not do it.

Where a response-scale contrast is wanted -- a natural direct/indirect effect on
the risk difference or risk-ratio scale -- the mediation formula must be applied
to the fitted models, which is outside missingmed's orchestration remit. That is
a fitting-layer concern (medfit), not a pooling one.

## 5A.2 IPW plus a binomial family: the non-integer-successes warning

`stats::glm()` emits `non-integer #successes in a binomial glm!` whenever a
binomial fit receives non-integer prior weights. IPW weights are inverse
sampling probabilities, not trial counts, so the warning fires on every
`method = "ipw"` fit with a binomial family. It is a false alarm: the weighted
score equations being solved are exactly the ones IPW specifies, and the point
estimates and sandwich SEs are unaffected. It is left unsuppressed rather than
silenced, so that a genuine non-integer-response mistake still surfaces.

---

# 5B. MNAR sensitivity (`sensitivity_mnar()`)

The pipeline estimates under **MAR** — `mice` imputes under MAR by construction.
MAR versus MNAR is not testable from the observed data, so there is nothing to
estimate without an assumption imported from outside it. `sensitivity_mnar()`
therefore produces a **curve**, not an estimate:

> if the unobserved values of the target sit `delta` units away from what MAR
> imputation implies, the indirect effect is *X*.

No rung is "the MNAR estimate". Each is conditional on its own delta.

## 5B.1 Method

Delta-adjusted imputation in the pattern-mixture sense (van Buuren, *FIMD* §9.2;
Leacy et al. 2017). Per rung: re-impute with the shift applied, then run the
existing pipeline unchanged.

For a **continuous** target the canonical procedure imputes under MAR and then
adds the constant to the imputed values (Hayati Rezvan et al. 2018) — which is
what `mice`'s `post` argument does, and what this implements. For a
**categorical** target the shift instead belongs on the imputation model's
linear predictor, with delta on the odds-ratio scale; that is not implemented,
and such targets error rather than receiving a meaningless additive shift.

Re-imputation rebuilds from the `mids` object's **stored settings**, never from
its recorded `call` — a `mids` built inside a function stores a call referencing
that function's local variables, which does not resolve anywhere else.

```r
sens <- sensitivity_mnar(md, delta = c(0, -0.5, -1, -1.5, -2))
tidy(sens)
summary(sens)   # adds the tipping point, if the grid contains one
```

## 5B.2 What `delta` actually is — read before choosing a value

`delta` is a **conditional** sensitivity parameter (CSP): a difference
conditional on all remaining variables *and their missingness indicators*. The
quantity an analyst can actually reason about — "non-respondents average half a
point lower" — is a **marginal** one (MSP). They are different numbers.

Supplying an elicited MSP where a CSP is expected is the standard failure mode
of this method. Tompsett et al. (2018) measure it: coverage fell from 95% to
**49.3%** when an elicited MSP was inserted directly as the CSP.

`missingmed` cannot make the specified parameter marginal, but it can show you
the realized one. Every rung reports `msp`:

| delta (CSP) | msp (realized) | estimate |
|---|---|---|
| 0 | 0.244 | 0.323 |
| −0.5 | −0.256 | 0.278 |
| −2 | −1.76 | 0.098 |

Two things to read off that table. First, `msp` is **non-zero at delta = 0** —
under MAR, imputed values legitimately differ from observed ones when
missingness depends on covariates. Second, `msp` is not simply `delta` shifted:
the gap is a chained-equations effect. With a single incomplete variable the
increments coincide exactly; with more, part of the shift circulates back
through the other imputation models.

## 5B.3 Reading a tipping point

`summary()` reports the delta at which the interval first includes zero. Judge
plausibility on the **realized `msp`**, not on the CSP: Tompsett et al. note a
CSP-scale tipping point has no direct clinical interpretation. Resseguier et al.
add the constraint that a tipping value must "correspond to reasonable
hypotheses supported by epidemiologic evidences" — a tipping point that requires
an implausible departure from MAR is *reassurance*, not a warning.

The figure below uses the same illustrative grid as the table in 5B.2. It shows
why the resolution warning matters: the interval's lower bound crosses zero near
**-1.6**, but a grid of `c(0, -0.5, -1, -1.5, -2)` can only *report* **-2.0**.

```{r sensitivity-curve, eval = TRUE, echo = FALSE, fig.width = 7, fig.height = 4.2, fig.alt = "Sensitivity curve of the indirect effect against delta with a confidence band; the reported tipping point at delta = -2 is marked alongside the interpolated true crossing near -1.6."}
d <- c(0, -0.5, -1, -1.5, -2)
est <- c(0.323, 0.278, 0.221, 0.160, 0.098)
se <- rep(0.075, 5)
lo <- est - 1.96 * se; hi <- est + 1.96 * se
par(mar = c(4.2, 4.2, 2.6, 1), mgp = c(2.4, 0.7, 0))
plot(d, est, type = "b", pch = 19, lwd = 2.2, col = "#4a76a8", ylim = c(-0.10, 0.52),
  xlab = "delta  (conditional sensitivity parameter)", ylab = "indirect effect  a*b",
  main = "A grid can only report a tipping point it landed on")
polygon(c(d, rev(d)), c(lo, rev(hi)), col = adjustcolor("#4a76a8", 0.15), border = NA)
lines(d, lo, col = "#4a76a8", lty = 3)
abline(h = 0, lwd = 1.8, col = "#b34a4a")
rep_tip <- d[which(lo <= 0)[1]]
true_tip <- approx(lo, d, xout = 0)$y
abline(v = rep_tip, lty = 2, col = "grey25")
abline(v = true_tip, lty = 2, col = "#b34a4a")
points(true_tip, 0, pch = 4, col = "#b34a4a", cex = 1.5, lwd = 2)
text(rep_tip, 0.47, sprintf("reported\ndelta = %.1f", rep_tip), pos = 4, cex = 0.78, col = "grey20")
text(true_tip, 0.30, sprintf("true crossing\n~ %.2f", true_tip), pos = 4, cex = 0.78, col = "#b34a4a")
legend("bottomright", "CI lower bound", lty = 3, col = "#4a76a8", bty = "n", cex = 0.78)
```

## 5B.4 Limitations

| Limitation | Status |
|---|---|
| `method = "ipw"` | Errors. IPW has no imputations to shift; a weighting analogue is a different method with a different parameter. |
| Categorical targets | Errors. Needs a linear-predictor offset (§5B.1); v2. |
| `pmm` targets | Allowed, with a message: shifted values may leave the observed range `pmm` otherwise guarantees. |
| Imputation-model compatibility | **Assumed, not checked.** The `mids` arrives pre-built; if its imputation model is incompatible with the mediation model, the curve confounds departure-from-MAR with misspecification. `smcfcs` is the upstream fix. |
| CSP calibration | Not implemented. Searching for the CSP that realizes a target MSP (Tompsett et al. §8) is the principled fix for §5B.2; v2. |

---

# 6. Cross-package engineering decisions

These changes were made upstream to satisfy missingmed's contracts; each is the
"small upstream fix" that a new capability turned out to need.

| Decision | Where | Why |
|---|---|---|
| `fit_mediation(weights=)` via `do.call` (value inlined), added only when non-NULL | medfit | `glm` evaluates `weights` by NSE in the **formula's** environment; passing it through `...` resolves to `stats::weights` (a function). `do.call` inlines the vector; gating on non-NULL keeps the unweighted path byte-identical. |
| `se_type` -> injectable `vcov_fun` in `extract_mediation` | medfit | Keeps named-vcov assembly in one place; `sandwich::vcovHC` swaps in for `stats::vcov` without missingmed re-implementing medfit's alias expansion. |
| Tidiers exported with `@exportS3Method broom::tidy` | missingmed | Plain `@export` registers `S3method(tidy, *)` against the wrong generic; the package-qualified form binds to broom's generic so `tidy()`/`broom::tidy()` dispatch. |
| `S7::S4_register()` per S7 class | missingmed | The legacy S4 generics (`print`/`summary`) require S7 classes to be S4-registered before S7 methods can attach. |
| `namespace` roclet enabled | missingmed | Lets roxygen regenerate NAMESPACE for the many new S7 exports (surfaced + fixed a latent `generics`/`broom` `tidy` import clash). |
| Pooled `MediationData` via copy-modify | missingmed | Inherits all medfit metadata; only estimates/vcov/paths change, so names stay intact for RMediation. |

---

# 7. Design-choice summary

- **S7-first**, three classes mirroring the S4 ancestors; estimator and model as
  orthogonal axes.
- **Delegate fitting to medfit** (`fit_mediation`) rather than re-implement GLM
  mediation — keeps missingmed thin and the `MediationData` naming canonical.
- **Names are the API**: pooled estimates/vcov stay named so RMediation resolves
  paths by label.
- **MBCO hosted locally** until RMediation offers an MI entry point; exact parity
  with the research prototype is the acceptance bar.
- **IPW = thin `run()` branch + passthrough `pool()`**; weights support both
  joint and sequential models, stabilized and trimmed; sandwich SE by default.
- **S4 deprecated** with `.Deprecated()` shims for one cycle.

---

# 7A. Self-check

*Four questions, about 10 minutes. Attempt each before opening its answer --
retrieval is what makes it stick.*

**1. Procedural** *(1 min, arithmetic)*
With \(m = 5\), \(\bar U = 0.04\) and \(B = 0.02\), compute \(T\) and the
fraction of missing information.

<details>
<summary>Answer</summary>

\(T = \bar U + (1+\tfrac15)B = 0.04 + 1.2(0.02) = 0.04 + 0.024 =
\mathbf{0.064}\).

\(\gamma = 0.024/0.064 = \mathbf{0.375}\) -- 37.5% of the information about
this parameter is missing.

*Stuck?* Re-read the worked example in section 3, steps 2 and 4.

</details>

**2. Conceptual** *(2 min)*
Section 4.1 shows \(T = 2[\ell_{\text{full}} - \max(\ell_{a=0},
\ell_{b=0})]\). Why does the `max()` -- rather than the product \(ab\) itself
-- prevent pooling before testing?

<details>
<summary>Answer</summary>

Rubin's rules are a **linear** operation: \(\bar Q\) and \(T\) are averages
of per-imputation quantities. `max()` is not linear, so the \(\max\) of the
pooled log-likelihoods is not the pool of the per-imputation \(\max\)es -- the
constrained fit may select the \(a=0\) branch in one imputation and the
\(b=0\) branch in another. That is why `MDMediationFit` retains the
per-imputation list.

The sharpest part: the product \(ab\) is *also* nonlinear, yet it causes no
such problem, because `pool()` pools \((a,b)\) linearly and defers the product
to `RMediation`. **A deferral is available for a product; none is available for a
branch selection.**

*Stuck?* Sections 4.1 and 4.2.

</details>

**3. Critical** *(3 min)*
A colleague reports `sensitivity_mnar(delta = c(0, -1, -2))` with a tipping point
at \(\delta = -2\), and concludes: *"the mediation effect is robust unless
non-respondents score 2 points lower."* Name two things wrong with that sentence.

<details>
<summary>Answer</summary>

1. **Wrong parameter.** \(\delta\) is *conditional*; "non-respondents score 2
   points lower" is a *marginal* claim. The realized `msp` at \(\delta = -2\)
   will not be \(-2\) -- read it off the output instead of assuming it
   (section 5B.2). Tompsett et al. (2018) measured the cost of this exact
   confusion: coverage fell from 95% to 49.3%.
2. **Wrong resolution.** Three grid points locate a tipping point no better than
   the grid spacing. The true crossing lies somewhere in \((-1, -2]\).

A third, if you want it: *"robust"* is a claim about **plausibility**, and
nothing in the output speaks to whether \(\delta = -2\) is plausible for this
study. The package can tell you where the conclusion flips; only you can say
whether that is a lot.

*Stuck?* Sections 5B.2 and 5B.3.

</details>

**4. Transfer** *(4 min -- nothing in this vignette answers it directly)*
You have a binary mediator with 30% missingness and want an MNAR sensitivity
curve. `sensitivity_mnar()` errors. What is the obstacle, and what would a
correct construction shift?

<details>
<summary>Answer</summary>

**The obstacle.** The implementation adds \(\delta\) to the **drawn value**
through `mice`'s `post`. Adding a constant to a drawn 0 or 1 produces something
that is neither -- so the package refuses rather than returning a meaningless
number.

**The correct construction.** Shift the imputation model's **linear predictor**,
which puts \(\delta\) on the log-odds scale. Concretely, a log-odds
\(\delta = 1\) moves an imputed prevalence of about 0.55 to about 0.76.

`mice` ships this as `mice.impute.mnar.logreg()` via `blots`; see
`docs/specs/SPEC-narfcs-delegation-2026-08-29.md` for the routing design.

*Stuck?* Section 5B.1's paragraph on categorical targets states the obstacle;
the construction is the part you had to supply.

</details>

## 7A.1 Three common confusions

**"`delta = 0` should give `msp = 0`."**
No. Under MAR, imputed values legitimately differ from observed ones whenever
missingness depends on covariates, because the imputed subgroup has a different
covariate mix. What \(\delta = 0\) guarantees is that the *analysis*
reproduces the MAR result exactly -- not that the marginal shift is zero. The
table in section 5B.2 shows `msp = 0.244` at `delta = 0`.

**"`se_type = "sandwich"` corrects for having estimated the weights."**
No. The sandwich is heteroskedasticity-consistent and treats \(\hat\pi\) as
fixed. Robins, Rotnitzky and Zhao (1994) show that ignoring weight estimation
makes the variance **conservative** -- estimating the weights *reduces* variance
relative to knowing them. The intervals are honest, just wider than necessary.

**"IPW and MI should agree; if they disagree, one is broken."**
No. They target the same estimand under MAR but use different information: MI
uses partially observed rows, IPW discards them and reweights what remains.
Disagreement beyond Monte-Carlo noise is **diagnostic** -- usually of a weight
model missing a MAR driver, or an imputation model incompatible with the
substantive model -- not of a bug.

---

# 8. References

- Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys*. Wiley.
- Chan, K. W., & Meng, X.-L. (2022). Multiple improvements of multiple
  imputation likelihood ratio tests. *Statistica Sinica*.
- Grund, S., Lüdtke, O., & Robitzsch, A. (2021). Pooling methods for
  likelihood-ratio tests with multiply imputed data. *Psychological Methods*.
- Seaman, S. R., & White, I. R. (2013). Review of inverse probability weighting
  for dealing with missing data. *Statistical Methods in Medical Research*.
- van Buuren, S. *Flexible Imputation of Missing Data*, 2nd ed., §9.2 (delta
  adjustment).
- Leacy, F. P., Floyd, S., Yates, T. A., & White, I. R. (2017). Analyses of
  sensitivity to the missing-at-random assumption using multiple imputation with
  delta adjustment. *American Journal of Epidemiology*, 185(4), 304-315.
- Tompsett, D. M., Leacy, F., Moreno-Betancur, M., Heron, J., & White, I. R.
  (2018). On the use of the not-at-random fully conditional specification
  (NARFCS) procedure in practice. *Statistics in Medicine*, 37(15), 2338-2353.
- Hayati Rezvan, P., Lee, K. J., & Simpson, J. A. (2018). Sensitivity analysis
  within the multiple imputation framework using delta-adjustment.
  *Longitudinal and Life Course Studies*, 9(3), 259-278.
- Resseguier, N., Giorgi, R., & Paoletti, X. (2011). Sensitivity analysis when
  data are missing not-at-random. *Epidemiology*, 22(2), 282-287.
