---
title: "A worked analysis with a missing mediator"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A worked analysis with a missing mediator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_deps <- requireNamespace("mice", quietly = TRUE) &&
  requireNamespace("medfit", quietly = TRUE) &&
  requireNamespace("RMediation", quietly = TRUE)
knitr::opts_chunk$set(eval = has_deps)
```

This tutorial walks through a complete mediation analysis when the mediator
is partly missing. It follows the steps of the illustrative analysis in the
methods paper behind missingmed, on simulated data so that every number here
can be reproduced from this page:

1. analyze the complete data, as a reference;
2. delete part of the mediator at random, given observed variables (MAR);
3. impute the mediator;
4. pool a Monte Carlo confidence interval for the indirect effect;
5. test the indirect effect with D4-MBCO;
6. repeat steps 3 to 5 on a dataset where both paths are weak;
7. ask how far the conclusion depends on MAR.

The tutorial has two parts. **Part 1** is the core analysis, which works the
same way in missingmed 0.5.0. **Part 2** shows what the development version
adds: checks that catch a mis-specified model before it is fitted, clearer
errors, a treatment-by-mediator interaction, and more control over the
sensitivity analysis. Each Part 2 section says what 0.5.0 did instead.

```{r, message = FALSE}
library(missingmed)
```

We call `mice::mice()` with its package prefix instead of attaching mice.
Both packages export a `pool()`; if you run `library(mice)` after
`library(missingmed)`, `pool(fit)` calls `mice::pool()` and fails with
"Argument 'object' not a list". Attach mice first, or use `missingmed::pool()`.

# Part 1: the core analysis

## The data

A randomized two-arm program (`treat`, 0 or 1) is meant to change an outcome
`Y` through a mediator `M`. Two baseline covariates, `base` and `hard`, are
measured before randomization and predict both `M` and `Y`. The function
below simulates such a study; the true paths are `a` (treatment to mediator),
`b` (mediator to outcome) and `cp` (the direct effect).

```{r}
simulate_study <- function(n, a, b, cp, seed) {
  set.seed(seed)
  base <- rnorm(n)
  hard <- rnorm(n)
  treat <- rbinom(n, 1, 0.5)
  M <- a * treat + 0.30 * base - 0.20 * hard + rnorm(n)
  Y <- b * M + cp * treat + 0.40 * base + 0.25 * hard + rnorm(n)
  data.frame(treat, base, hard, M, Y)
}

full <- simulate_study(n = 600, a = 0.35, b = 0.30, cp = 0.10, seed = 4021)
head(full)
```

The true indirect effect is `a * b = 0.35 * 0.30 = 0.105`.

## Step 1: the complete-data reference

Before deleting anything, fit the two regressions on the full data. missingmed
fits through `medfit`, so the same call works on any data frame; here we use
`medfit::fit_mediation()` directly.

```{r}
ref <- medfit::fit_mediation(
  formula_y = Y ~ treat + M + base + hard,
  formula_m = M ~ treat + base + hard,
  data = full, treatment = "treat", mediator = "M"
)
c(a = ref@a_path, b = ref@b_path, ab = ref@a_path * ref@b_path)
```

The estimates are close to the truth. This is the answer the analyses below
should recover.

## Step 2: make the mediator missing at random

The mediator is deleted with a probability that rises with baseline `base`,
with `hard`, and in the treated arm. Each of those is observed, so the
mediator is missing at random (MAR): given the observed variables, whether
`M` is missing does not depend on the value of `M` itself.

```{r}
set.seed(77)
p_miss <- plogis(-1.4 + 0.6 * full$base + 0.4 * full$treat + 0.3 * full$hard)
obs <- full
obs$M[runif(nrow(obs)) < p_miss] <- NA
mean(is.na(obs$M))
```

About a quarter of the mediator values are now missing. Dropping those rows
(complete-case analysis) would lose that much data and, because the rows lost
are not a random subset, can bias the estimates.

## Step 3: impute the mediator

Multiple imputation replaces each missing value with several plausible ones.
Two choices matter:

- **Include the outcome in the imputation model.** `mice` uses every other
  column as a predictor by default, so `Y` helps impute `M`. Leaving it out
  would make the imputation model assume no relation between `M` and `Y`, and
  pull the `b` path toward zero.
- **Use enough imputations.** We use `m = 10`; more imputations reduce the
  Monte Carlo noise in the pooled results.

```{r}
imp <- mice::mice(obs, m = 10, method = "pmm", printFlag = FALSE, seed = 2718)
```

Now describe the mediation model once. `set_md_mediation()` stores the data,
the two regressions and the roles of the variables; nothing is fitted yet.

```{r}
md <- set_md_mediation(imp,
  formula_y = Y ~ treat + M + base + hard,
  formula_m = M ~ treat + base + hard,
  treatment = "treat", mediator = "M"
)
md
```

`run()` fits both regressions on each of the 10 completed datasets, and
`pool()` combines them with Rubin's rules.

```{r}
fit <- run(md)
res <- pool(fit)
res
```

Rubin's rules average the 10 estimates, and add the spread between
imputations to the average within-imputation variance. That extra spread is
the price of not observing the missing values.

## Step 4: a pooled Monte Carlo interval

The indirect effect is a product, `a * b`, and its sampling distribution is
not normal. `infer(type = "mc")` draws `(a, b)` from a normal distribution
with the pooled estimates and pooled covariance, and reads the interval off
the distribution of the products.

```{r}
set.seed(1)
mc <- infer(fit, type = "mc")
mc$CI
```

The interval excludes zero and covers the true value 0.105. Note that
`mc$Estimate` is the mean of the draws, which includes the covariance of `a`
and `b`; it is not exactly the product of the pooled `a` and `b`.

## Step 5: test the indirect effect with D4-MBCO

The Monte Carlo interval answers "which values of `a * b` are plausible". A
test answers "is `a * b` zero". The model-based constrained optimization
(MBCO) test compares the fit of the full model with the best fit in which the
indirect effect is zero, that is, with `a = 0` or with `b = 0`.

Under multiple imputation, MBCO cannot simply be averaged across imputations.
`infer(type = "mbco")` uses the D4 statistic instead: it fits the full and
the constrained models to all imputed datasets stacked together, then
corrects for the extra variance due to imputation.

```{r}
mbco <- infer(fit, type = "mbco", ariv = "fixed")
mbco
mbco[["p"]]
```

The printout reports the D4 statistic and its F reference distribution, the
relative increase in variance due to missing data (`r4`), and which path the
stacked constrained fit set to zero. Here every imputation agrees that
setting `a = 0` is the better constrained fit, so the two `ariv` choices
explained in step 6 give the same answer.

## Step 6: when both paths are weak

The same steps on a study with small `a` and `b`:

```{r}
weak <- simulate_study(n = 600, a = 0.10, b = 0.09, cp = 0.05, seed = 41)
set.seed(78)
p_miss <- plogis(-1.4 + 0.6 * weak$base + 0.4 * weak$treat + 0.3 * weak$hard)
weak$M[runif(nrow(weak)) < p_miss] <- NA

imp_w <- mice::mice(weak, m = 10, method = "pmm", printFlag = FALSE, seed = 2718)
fit_w <- run(set_md_mediation(imp_w,
  formula_y = Y ~ treat + M + base + hard,
  formula_m = M ~ treat + base + hard,
  treatment = "treat", mediator = "M"
))
set.seed(2)
infer(fit_w, type = "mc")$CI
```

With both paths near zero, the imputations can disagree about which path to
constrain: in some completed datasets the `a = 0` model fits better, in
others the `b = 0` model does. The `ariv` argument says how D4 handles that.

```{r}
fixed <- infer(fit_w, type = "mbco", ariv = "fixed")
own <- infer(fit_w, type = "mbco", ariv = "own")
fixed
rbind(fixed = S7::S7_data(fixed), own = S7::S7_data(own))
```

- `ariv = "fixed"` (the default) takes the branch the **stacked** constrained
  fit chose and tests every imputation on that same branch.
- `ariv = "own"` lets each imputation use its own better branch when computing
  the variance correction `r4`.

When the branches are mixed, as here (`branch_mix` is `TRUE` in the printout),
the two choices give different `r4` and p-values. When they are not mixed, as
in step 5, they agree exactly. Report which one you used; `mbco_d4()`
documents both.

## Step 7: how much does the conclusion depend on MAR?

MAR cannot be checked from the observed data. `sensitivity_mnar()` asks what
happens if the missing mediator values are systematically higher or lower than
MAR imputation assumes. It re-imputes with the imputed values shifted by
`delta` (in the mediator's units) and repeats the inference at each shift.

```{r, message = FALSE}
sens <- sensitivity_mnar(md, delta = c(0, -0.5, -1, -1.5, -2),
  type = "mc", n.mc = 2e4
)
summary(sens)
```

`delta = 0` is the MAR analysis. The tipping point is the smallest shift at
which the interval includes zero. Column `msp` is the mean of the imputed
mediator values minus the mean of the observed ones. It is not zero even at
`delta = 0`: under MAR the people with a missing mediator have higher `base`,
so their imputed values are higher on average. Compare each row's `msp` with
the `delta = 0` row to see how much the shift moved the imputations, and judge
whether a departure that large is plausible for your study.

# Part 2: what the development version adds

Everything in Part 1 runs the same way in missingmed 0.5.0. The sections
below show behavior that is new in the development version, and what 0.5.0
did instead.

## The model is checked before it is fitted

A typo in a formula used to be fitted without complaint. Here the mediator
model regresses a covariate, not the mediator, on the treatment:

```{r, error = TRUE}
set_md_mediation(imp,
  formula_y = Y ~ treat + M + base + hard,
  formula_m = base ~ treat + hard,
  treatment = "treat", mediator = "M"
)
```

In 0.5.0 this call succeeded, and the pipeline reported the effect of
treatment on `base` as the `a` path, a silently wrong indirect effect.
`set_md_mediation()` now also refuses, with a message naming the problem:

- a treatment or mediator missing from either formula;
- a product or transform involving the treatment or mediator other than one
  `treat:M` term (`mbco_d4()` accepts moderated models for testing);
- a non-numeric treatment (recode a factor to 0/1);
- an engine missingmed cannot run.

```{r, error = TRUE}
set_md_mediation(imp,
  formula_y = Y ~ treat + M + base + hard,
  formula_m = M ~ treat + base + hard,
  treatment = "treat", mediator = "M", engine = "lavaan"
)
```

In 0.5.0 an unsupported engine failed later, inside `run()`, with an error
from medfit. When a fit does fail, the error now names the engine and the
imputation (`engine "glm" failed on imputation 3 of 10: ...`), and fitting
warnings are collected into one warning that lists the affected imputations.

## Arguments that do not apply are reported

```{r}
r <- infer(fit, type = "mbco", level = 0.90)
```

MBCO is a test, so it has no confidence level. In 0.5.0, `level` (and
`n.mc`, `treatment_level`) were silently ignored under `type = "mbco"`, and a
misspelled argument such as `conf.level = 0.9` was silently dropped from a
`"mc"` call, returning a 95% interval. Now the first case warns and the second
is an error:

```{r, error = TRUE}
infer(fit, type = "mc", conf.level = 0.90)
```

## A treatment-by-mediator interaction

If the effect of the mediator on the outcome differs between arms, add
`treat:M` to the outcome model. The indirect effect is then
`a * (b + theta3 * x)`: it depends on the treatment level `x` at which the
mediator's effect is evaluated.

```{r}
fit_xm <- run(set_md_mediation(imp,
  formula_y = Y ~ treat * M + base + hard,
  formula_m = M ~ treat + base + hard,
  treatment = "treat", mediator = "M"
))
set.seed(3)
infer(fit_xm, type = "mc", treatment_level = 1)$CI
set.seed(3)
infer(fit_xm, type = "mc", treatment_level = 0)$CI
```

For a 0/1 treatment, `treatment_level = 1` gives the total natural indirect
effect and `0` the pure natural indirect effect. In these data there is no
true interaction, so the two intervals overlap heavily. In 0.5.0, `pool()` failed
on such a model; `treatment_level` is required here so the estimand is always
stated.

## Sensitivity analysis with the MBCO test

`sensitivity_mnar()` can run the D4-MBCO test at every shift, and now passes
`ariv` through (0.5.0 always used `"fixed"`):

```{r, message = FALSE}
sens_mbco <- sensitivity_mnar(md, delta = c(0, -1, -2),
  type = "mbco", ariv = "fixed"
)
sens_mbco
summary(sens_mbco)
```

For an MBCO curve the null is retained at a shift when `p > 1 - level`. If a
rung fails to produce a result (an `NA` row), `summary()` still reports the
tipping point when every such rung lies farther from MAR than it, because
those rungs cannot change the answer; otherwise it says the tipping point is
undetermined and names the rungs. In 0.5.0, any `NA` rung made `summary()`
fail.

# Summary

| Step | Function | Result |
|---|---|---|
| Describe the model | `set_md_mediation()` | checked roles and formulas |
| Fit each imputation | `run()` | per-imputation fits |
| Pool | `pool()` | Rubin's-rules estimates |
| Interval | `infer(type = "mc")` | Monte Carlo CI for the indirect effect |
| Test | `infer(type = "mbco")` | D4-MBCO test of `a * b = 0` |
| Departures from MAR | `sensitivity_mnar()` | a sensitivity curve and tipping point |

See `vignette("mbco-mi")` for more on the D4-MBCO test and
`vignette("technical")` for the pooling details.
