---
title: "S7 classes and methods"
format:
  html:
    toc: true
    code-copy: hover
execute:
  eval: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(missingmed)
```

A developer's map of missingmed's object model: which S7 class each verb takes
and returns, what each class carries, and which generics dispatch on it. For
the statistics behind the verbs, see `vignette("technical")`.

# The pipeline

Four verbs move a mediation analysis through three classes. Inference returns
either an `RMediation` interval or a fourth class, and the sensitivity analysis
returns a fifth.

```{mermaid}
flowchart LR
  D["MDMediationData"] -->|"run()"| F["MDMediationFit"]
  F -->|"pool()"| R["MDMediationResult"]
  R -->|"infer(type = 'mc')"| CI["RMediation interval (list)"]
  F -->|"infer(type = 'mc')"| CI
  F -->|"infer(type = 'mbco')"| M["MbcoMIResult"]
  D -->|"sensitivity_mnar()"| S["MDSensitivityResult"]
  S -. "one infer() result per rung" .-> CI
  S -.-> M
```

`infer(type = "mbco")` takes the **fit**, not the pooled result: the MBCO
statistic does not commute with Rubin's rules, so it needs every imputation.
`infer(<MDMediationResult>, type = "mbco")` is an error by design.

# Classes

Each class below is an `S7::new_class()` with `package = "missingmed"`. The
property lists are read from the class objects when this page is built.

```{r}
#| label: props
#| echo: false
classes <- list(
  MDMediationData = MDMediationData,
  MDMediationFit = MDMediationFit,
  MDMediationResult = MDMediationResult,
  MbcoMIResult = MbcoMIResult,
  MDSensitivityResult = MDSensitivityResult
)
props <- data.frame(
  class = names(classes),
  parent = vapply(classes, function(k) {
    p <- k@parent
    if (inherits(p, "S7_base_class")) paste0("class_", p$class) else p@name
  }, ""),
  properties = vapply(classes, function(k) {
    paste0("`", names(k@properties), "`", collapse = ", ")
  }, ""),
  row.names = NULL
)
knitr::kable(props)
```

- **`MDMediationData`**, built by `set_md_mediation()`, holds the analysis
  specification: the data (a `mice::mids` for `method = "mi"`, a data frame for
  `method = "ipw"`), the two model formulas and families, the treatment and
  mediator names, and the IPW weight settings.
- **`MDMediationFit`**, returned by `run()`, holds one named
  `medfit::MediationData` per imputation in `per_imputation`, plus the raw fits
  and, for IPW, the weights. `source` points back to the `MDMediationData`,
  which is how `infer(type = "mbco")` reaches the imputed datasets.
- **`MDMediationResult`**, returned by `pool()`, holds the Rubin's-rules pooled
  `medfit::MediationData` in `pooled`, its tidy table with per-coefficient Wald
  tests, and the within, between and total covariance matrices.
- **`MbcoMIResult`**, returned by `infer(type = "mbco")` and `mbco_d4()`, is
  the only class with a base-type parent (`class_double`). Its data is the
  named vector `c(D4, p, r4, nu, d_S)`; its properties record `ariv`, the
  numerator df `k`, the number of imputations `m`, and the branch diagnostics.
- **`MDSensitivityResult`**, returned by `sensitivity_mnar()`, holds one
  inference result per `delta` rung in `rungs`, the `grid` of deltas, and the
  realized marginal sensitivity parameter `msp` for each rung.

# Generics and methods

```{r}
#| label: methods
#| echo: false
gens <- list(
  run = run, pool = pool, infer = infer,
  per_imputation_list = per_imputation_list, n_imputations = n_imputations
)
dispatch <- data.frame(
  generic = names(gens),
  `methods for` = vapply(gens, function(g) {
    paste0("`", sort(sub("^missingmed::", "", names(g@methods))), "`",
      collapse = ", "
    )
  }, ""),
  check.names = FALSE, row.names = NULL
)
knitr::kable(dispatch)
```

The methods on the wrong class are refusals with a pointer to the right step:
`pool(<MDMediationData>)` asks for `run()` first, and `pool(<MDMediationResult>)`
says the object is already pooled. `pool()` on any other object falls through
to `mice::pool()`.

Output methods are S7 methods on external generics:

| Generic | Classes |
|---|---|
| `print()` | all five |
| `summary()` | `MDMediationResult`, `MDSensitivityResult` |
| `broom::tidy()` (re-exported) | `MDMediationResult`, `MbcoMIResult`, `MDSensitivityResult` |
| `[`, `[[` | `MbcoMIResult` (index the underlying vector) |

# Working with an `MbcoMIResult`

S7 objects are not subsettable by default, so `MbcoMIResult` defines `[` and
`[[` on its data. Properties use `@`; `$` is an error.

```{r}
#| label: mbco
set.seed(1)
implist <- lapply(1:3, function(i) {
  X <- rnorm(200)
  M <- 0.4 * X + rnorm(200)
  data.frame(X = X, M = M, Y = 0.3 * M + rnorm(200))
})
r <- mbco_d4(implist, Y ~ X + M, M ~ X,
  treatment = "X", mediator = "M", ariv = "fixed"
)
r[["p"]]
r[c("D4", "nu")]
r@stacked_branch
is.numeric(r)
S7::S7_data(r)
```

# Registration notes for contributors

- `R/zzz.R` calls `S7::methods_register()` in `.onLoad()`, which registers the
  S7 methods on external generics at load time, so they need no `S3method()`
  lines in `NAMESPACE`.
- `MDMediationData`, `MDMediationFit`, `MDMediationResult` and
  `MDSensitivityResult` are passed to `S7::S4_register()`, so S4 code can
  dispatch on them.
- `MbcoMIResult` is **not**: `setOldClass()` cannot build an S4 prototype for
  an S7 class whose parent is `class_double`. Two consequences:
  - S7 refuses to add a method for an unregistered class to an S4 generic,
    and `print` can be one inside the namespace (it was made one by the S4
    classes and the `OpenMx` import that 0.6.0 removed). Its print method is
    therefore registered on `base::print`, which works either way.
  - Register methods for it with the functional form,
    `` S7::`method<-`(generic, MbcoMIResult, value = f) ``. The assignment form
    `S7::method(generic, MbcoMIResult) <- f` assigns the generic back into the
    namespace; for `[` and `[[` that creates objects `R CMD check` reports as
    undocumented.
- Every exported class and function needs an entry in the `reference:` index
  of `_pkgdown.yml`; `R CMD check` does not read it, but the pkgdown CI job
  fails without it.

# Removed S4 classes

The S4 API (`set_sem()`, `run_sem()`, `pool_sem()` with the classes
`SemImputedData`, `SemResults` and `PooledSEMResults`) was removed in 0.6.0; the
functions were `.Defunct()` stubs in 0.6.0 and were deleted in 0.7.0. Each S7 class above replaced one
of the classes:

| S7 class | Replaces |
|---|---|
| `MDMediationData` | `SemImputedData` |
| `MDMediationFit` | `SemResults` |
| `MDMediationResult` | `PooledSEMResults` |
