---
title: "Frequently asked questions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

### `pool()` gives an error about `mira` objects, or does the wrong thing

`mice` also exports `pool()`. Whichever package is attached last masks the
other, so `library(mice)` after `library(missingmed)` makes `pool()` mean
`mice::pool()`. Call `mice::mice()` with the prefix instead of attaching
`mice`, or call `missingmed::pool()`. The missingmed version forwards anything
that is not a missingmed fit (a `mice::mira`, for example) to `mice::pool()`,
so it is safe to keep missingmed's `pool()` in front.

### Monte Carlo interval or MBCO test?

They answer different questions. `infer(type = "mc")` gives a confidence
interval for the indirect effect from the pooled estimates. `infer(type =
"mbco")` gives a likelihood-ratio test of \(H_0: ab = 0\) and needs the
per-imputation fits, so call it on the result of `run()`, not of `pool()`.
`vignette("mbco-mi")` covers the test; `vignette("worked-analysis")` uses both.

### Why does `infer()` ask for `treatment_level`?

The outcome model has a treatment-by-mediator interaction, so the indirect
effect is \(a (b + \theta_3 x)\) and depends on the treatment level \(x\).
missingmed does not pick \(x\) for you. For a 0/1 treatment, `1` gives the total
and `0` the pure natural indirect effect. Section 3A of
`vignette("technical")` has the details.

### What does `ariv` do?

It applies to `infer(type = "mbco")` and `mbco_d4()`. With `ariv = "fixed"`
(the default), every imputation's likelihood ratio is taken on the branch the
stacked constrained fit chose (\(a = 0\) or \(b = 0\)). With `ariv = "own"`,
each imputation uses its own winning branch, which reproduces missingmed 0.4.0.
The branch diagnostics are properties of the returned `MbcoMIResult`.

### How many imputations do I need?

The pooled variance adds \((1 + 1/m)B\) to the within-imputation variance, and
the Monte Carlo error of everything computed across imputations shrinks as
\(m\) grows. A practical check: rerun with a larger `m` and a different seed,
and see whether the conclusions move. The `fmi` column of
`pool(fit)@tidy_table` reports the fraction of missing information per
coefficient.

### My model has a moderator. Can I use it?

Not in `set_md_mediation()`, which allows only main effects of the treatment
and mediator plus one `X:M` term (see the *Supported models* article).
`mbco_d4()` tests the indirect effect of a moderated model on a list of
completed datasets.

### Binary mediator or outcome?

Set `family_m = binomial()` or `family_y = binomial()`. The indirect effect is
then a product of coefficients on the link scale, not a difference in
probabilities; section 5A of `vignette("technical")` explains the caveat.

### MI or IPW?

`method = "mi"` takes a `mids` object and pools across imputations. `method =
"ipw"` takes the raw data frame and reweights the complete cases by the inverse
probability of being complete, with sandwich standard errors by default. MBCO
is not available for IPW fits.

### What scale is `delta` in `sensitivity_mnar()` on?

The scale of the target variable's imputation model: an additive shift for a
continuous target, a log-odds shift for a binary one. Section 5B.2 of
`vignette("technical")` shows how to translate a delta into a statement about
non-respondents.

### Why does `infer()` warn that an argument was ignored?

Some arguments apply to one inference type only, such as `level`, `n.mc`
and `treatment_level` for `"mc"` and `ariv` for `"mbco"`. Passing one to the other
type warns, so a typo or a mismatched argument is not silently dropped.

### Structural equation models and latent mediators?

Yes, since 0.6.0. The glm path fits regression models through medfit; for a
structural equation model, including a latent mediator, use
`set_md_mediation(engine = "lavaan")` with the lavaan syntax in `model` and the
`outcome` named. `infer(type = "mc")` and `infer(type = "mbco")` both work for
lavaan fits (MBCO with maximum likelihood only; it is refused for `MLR`, `MLM`,
`WLSMV`, `group`, `ordered` and `sampling.weights`). See the *Structural equation
models* article and the *Migrating from the S4 API* article.

### How do I cite missingmed?

```{r}
citation("missingmed")
```
