---
title: "Inverse probability weighting, step by step"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Multiple imputation fills in the missing values. Inverse probability weighting
(IPW) does something different: it keeps only the **complete cases** and weights
each one by the inverse of its estimated probability of being observed, so the
complete cases stand in for everyone. missingmed runs both through the same
`set_md_mediation() -> run() -> pool() -> infer()` pipeline, selected by
`method`.

## Data with a missing mediator

Missingness in `M` depends on the covariate `C` (missing at random).

```{r data}
library(missingmed)
set.seed(1)
n <- 400
d <- data.frame(X = rbinom(n, 1, 0.5), C = rnorm(n))
d$M <- 0.5 * d$X + 0.3 * d$C + rnorm(n)
d$Y <- 0.3 * d$M + 0.2 * d$X + 0.3 * d$C + rnorm(n)
d$M[runif(n) < plogis(-1 + 0.8 * d$C)] <- NA
mean(is.na(d$M))
```

## Fit

For IPW, `data` is the raw data frame (not a `mids` object), and `method = "ipw"`.

```{r fit}
md <- set_md_mediation(d, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", method = "ipw"
)
fit <- run(md)
res <- pool(fit)
res
```

There is one fit (`m = 1`), so the pooled table is that fit's table, with robust
standard errors.

```{r infer}
set.seed(2)
infer(fit, type = "mc", n.mc = 20000)
```

## Look at the weights

`fit@weights` has one entry per row of the data: the weight for a complete case,
`NA` for an incomplete one.

```{r weights}
w <- fit@weights
c(complete = sum(!is.na(w)), incomplete = sum(is.na(w)))
summary(w)
```

A few very large weights dominate the fit. If the maximum is many times the
median, the missingness model has probabilities close to zero for some cases and
the estimate rests on them. Two remedies, both arguments of `set_md_mediation()`:
`weight_trim` caps the weights, and a simpler `weight_formula` avoids extreme
probabilities.

## Choosing the missingness model

By default the missingness model uses every observed predictor. `weight_formula`
overrides it: a single formula, or a named list with one formula per incomplete
variable.

```{r wf}
md2 <- set_md_mediation(d, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", method = "ipw",
  weight_formula = list(M = ~ X + C)
)
summary(run(md2)@weights)
```

**One model or one per variable.** The default, and a single formula, fit **one**
model for "every model variable is observed". A named list fits **one model per
incomplete variable, each on all rows,** and multiplies the probabilities.
Which is right depends on how the values go missing, which the data cannot show:

| Missingness | Joint model | Per-variable list |
| --- | --- | --- |
| One variable incomplete | correct | correct |
| Several incomplete, missing separately | biased | correct |
| Several incomplete, missing together | correct | biased |

"Biased" here means at any sample size. "Separately" means each variable's
missingness depends only on the observed predictors, not on the others'; "together"
means one event drops a row's values.

In the separate case the probability that all are observed is a *product* of
logistic probabilities, which one logistic model cannot represent. In the joint
case it equals each variable's own probability, so the product double-counts.
In a simulated check (`M` and `Y` each missing 40% of the time, driven by an
auxiliary variable, true `b` = 0.30), the joint default converged to `b` = 0.278
at n = 100,000 under separate missingness, while a per-variable
`weight_formula = list(M = ..., Y = ...)` converged to 0.300. In a pilot of the
opposite structure (laptop, 8 replications at n = 100,000, bias against the
full-data estimate), the per-variable list was about 0.05 too high and the joint
model unbiased. If several variables are incomplete, name every one of them in the
list (`run()` refuses a list that omits an incomplete variable).

## Stabilizing and trimming

`weight_stabilize = TRUE` (the default) multiplies by the marginal probability of
being observed, which leaves the estimate's target unchanged and shrinks the
weights' spread. `weight_trim` is an upper quantile: `0.95` caps the largest 5% of
weights at the 95th percentile. `1` (default) means no trimming.

```{r trim}
md3 <- set_md_mediation(d, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", method = "ipw", weight_trim = 0.95
)
rbind(
  untrimmed = summary(run(md)@weights)[c("Min.", "Median", "Max.")],
  trimmed   = summary(run(md3)@weights)[c("Min.", "Median", "Max.")]
)
```

Trimming trades a little bias for less variance. Report whether you trimmed. In
the simulated mechanism above (n = 200 or 500, 25% or 40% missing in each of `M`
and `Y`), trimming at 0.95 moved the mean `b` a further 0.02 to 0.03 below the
untrimmed stabilized weights (0.285 to 0.264, 0.282 to 0.258, 0.268 to 0.244 and
0.265 to 0.237 across the four settings).

## Standard errors

`se_type = "sandwich"` (default) uses a heteroskedasticity-robust (HC) variance;
`"model"` uses the usual model-based one, which is wrong for weighted fits and is
there for comparison only. The weights are treated as **known**: the uncertainty
from estimating them is not propagated. A variance that does propagate it exists
internally but is not enabled: in a 48-setting simulation (2,000 replications each)
it was no better than the current one.

Two things limit how far the intervals can be trusted, whatever the variance:

* **Misspecified weights** (the joint model above) bias the estimate at any sample
  size, so the interval is centered in the wrong place. In the worst simulated
  setting (n = 500, 40% missing in each variable) the 95% interval for `a*b` at
  `a = b = 0.3` covered 0.921 of the time with the joint default and 0.929 with
  per-variable weights.
* **Few complete cases.** Even with correct weights the estimate was biased low at
  small counts: with 40% missing in each of `M` and `Y` (about 36% complete cases),
  the mean `b` was 0.282 at n = 200, 0.288 at n = 500 and 0.301 at n = 5,000 (true
  value 0.30). Treat IPW results from a few hundred complete cases with heavy
  weights as approximate, and look at `summary(fit@weights)`.

## IPW or imputation?

Compare them on the same data.

```{r compare}
imp <- mice::mice(d, m = 10, method = "norm", printFlag = FALSE, seed = 3)
mi <- pool(run(set_md_mediation(imp, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M"
)))
ipw <- res
cmp <- function(r) setNames(r@tidy_table$estimate, r@tidy_table$term)[c("a", "b", "c_prime")]
round(rbind(MI = cmp(mi), IPW = cmp(ipw)), 3)
```

Both rely on the missing-at-random assumption. Imputation uses every row and
usually has smaller standard errors; IPW is simpler to describe and does not need
an imputation model for the mediator, but depends on the missingness model and
discards incomplete rows. When they disagree by more than their standard errors,
suspect a misspecified model on one side.

## Limits

* `sensitivity_mnar()` is not available for `method = "ipw"`: delta adjustment
  shifts imputed values, and IPW has none.
* `infer(type = "mbco")` is not implemented for IPW. Use `type = "mc"`.
* `engine = "lavaan"` works with IPW, with robust standard errors only; see the
  *Structural equation models* article.
