---
title: "Inverse probability weighting, step by step"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Multiple imputation fills in the missing values. Inverse probability weighting
(IPW) does something different: it keeps only the **complete cases** and weights
each one by the inverse of its estimated probability of being observed, so the
complete cases stand in for everyone. missingmed runs both through the same
`set_md_mediation() -> run() -> pool() -> infer()` pipeline, selected by
`method`.

## Data with a missing mediator

Missingness in `M` depends on the covariate `C` (missing at random).

```{r data}
library(missingmed)
set.seed(1)
n <- 400
d <- data.frame(X = rbinom(n, 1, 0.5), C = rnorm(n))
d$M <- 0.5 * d$X + 0.3 * d$C + rnorm(n)
d$Y <- 0.3 * d$M + 0.2 * d$X + 0.3 * d$C + rnorm(n)
d$M[runif(n) < plogis(-1 + 0.8 * d$C)] <- NA
mean(is.na(d$M))
```

## Fit

For IPW, `data` is the raw data frame (not a `mids` object), and `method = "ipw"`.

```{r fit}
md <- set_md_mediation(d, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", method = "ipw"
)
fit <- run(md)
res <- pool(fit)
res
```

There is one fit (`m = 1`), so the pooled table is that fit's table, with robust
standard errors.

```{r infer}
set.seed(2)
infer(fit, type = "mc", n.mc = 20000)
```

## Look at the weights

`fit@weights` has one entry per row of the data: the weight for a complete case,
`NA` for an incomplete one.

```{r weights}
w <- fit@weights
c(complete = sum(!is.na(w)), incomplete = sum(is.na(w)))
summary(w)
```

A few very large weights dominate the fit. If the maximum is many times the
median, the missingness model has probabilities close to zero for some cases and
the estimate rests on them. Two remedies, both arguments of `set_md_mediation()`:
`weight_trim` caps the weights, and a simpler `weight_formula` avoids extreme
probabilities.

## Choosing the missingness model

By default the missingness model is sequential (below) and uses every fully
observed model variable as a predictor. `weight_formula` overrides it: a single
formula, or a named list with one formula per incomplete variable.

```{r wf}
md2 <- set_md_mediation(d, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", method = "ipw",
  weight_formula = list(M = ~ X + C)
)
summary(run(md2)@weights)
```

**A sequence of models, or one.** A row is complete with probability
P(`M` observed) x P(`Y` observed given `M` observed), each factor given the
observed predictors. The default and a named list fit exactly that, each factor
on the rows where the earlier variables are observed. The default puts the
variable with fewer missing values first; a list is used in the order written.
A single formula instead fits **one** model for "every model variable is
observed". Which is right depends on how the values go missing, which the data
cannot show:

| Missingness | Default or list (sequential) | Single formula (joint) |
| --- | --- | --- |
| One variable incomplete | correct | correct |
| Several incomplete, separately | correct | biased |
| Several incomplete, together | correct | correct |
| Several incomplete, one after another | correct if ordered right | biased |

"Biased" here means at any sample size. "Separately" means each variable's
missingness depends only on the observed predictors, not on the others';
"together" means one event drops a row's values.

When the variables go missing separately the probability that all are observed
is a *product* of logistic probabilities, which one logistic model cannot
represent. When they go missing together, `Y` is observed on every row where `M`
is, so its factor is 1: the sequence reduces to one model and fits nothing for
`Y`. When one drops out after another, the order matters: the factor of the
variable that drops out second is the one that is logistic given the first, so
list the variables in the order in which they drop out if the default (fewest
missing first) is not that order. If several variables are incomplete, name
every one of them in a list (`run()` refuses a list that omits an incomplete
variable).

In a simulation with known weights (162 settings, n up to 5,000, 25% and 40%
missing in each variable), the largest bias in `b` at n = 5,000 was 0.004 for the
sequential model, 0.024 for a single formula (separate missingness) and 0.046
for the earlier per-variable list that fitted each variable on all rows (missing
together). At n = 500 or fewer, where the forms differ, the sequential model's
root mean squared error for `b` was 3% to 13% above the single formula's, even
where the single formula was biased; coverage differed by at most 0.007. The
bias that remains at a few hundred complete cases (below) is shared by every
form.

## Stabilizing and trimming

`weight_stabilize = TRUE` (the default) multiplies by the marginal probability of
being observed, which leaves the estimate's target unchanged and shrinks the
weights' spread. `weight_trim` is an upper quantile: `0.95` caps the largest 5% of
weights at the 95th percentile. `1` (default) means no trimming.

```{r trim}
md3 <- set_md_mediation(d, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", method = "ipw", weight_trim = 0.95
)
rbind(
  untrimmed = summary(run(md)@weights)[c("Min.", "Median", "Max.")],
  trimmed   = summary(run(md3)@weights)[c("Min.", "Median", "Max.")]
)
```

Trimming trades a little bias for less variance. Report whether you trimmed. In
the simulated mechanism above (n = 200 or 500, 25% or 40% missing in each of `M`
and `Y`), trimming at 0.95 moved the mean `b` a further 0.02 to 0.03 below the
untrimmed stabilized weights (0.285 to 0.264, 0.282 to 0.258, 0.268 to 0.244 and
0.265 to 0.237 across the four settings).

## Standard errors

`se_type = "sandwich"` (default) uses a heteroskedasticity-robust (HC) variance;
`"model"` uses the usual model-based one, which is wrong for weighted fits and is
there for comparison only. The weights are treated as **known**: the uncertainty
from estimating them is not propagated. A variance that does propagate it exists
internally but is not enabled: in a 48-setting simulation (2,000 replications each)
it was no better than the current one.

Two things limit how far the intervals can be trusted, whatever the variance:

* **Misspecified weights** (a single formula when the variables go missing
  separately, above) bias the estimate at any sample size, so the interval is
  centered in the wrong place. In the worst simulated setting (n = 500, 40%
  missing in each variable) the 95% interval for `a*b` at `a = b = 0.3` covered
  0.921 of the time with one joint model and 0.929 with per-variable weights,
  the form that is correct for that mechanism.
* **Few complete cases.** Even with correct weights the estimate was biased low at
  small counts: with 40% missing in each of `M` and `Y` (about 36% complete cases),
  the mean `b` was 0.282 at n = 200, 0.288 at n = 500 and 0.301 at n = 5,000 (true
  value 0.30). Treat IPW results from a few hundred complete cases with heavy
  weights as approximate, and look at `summary(fit@weights)`.

## IPW or imputation?

Compare them on the same data.

```{r compare}
imp <- mice::mice(d, m = 10, method = "norm", printFlag = FALSE, seed = 3)
mi <- pool(run(set_md_mediation(imp, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M"
)))
ipw <- res
cmp <- function(r) setNames(r@tidy_table$estimate, r@tidy_table$term)[c("a", "b", "c_prime")]
round(rbind(MI = cmp(mi), IPW = cmp(ipw)), 3)
```

Both rely on the missing-at-random assumption. Imputation uses every row and
usually has smaller standard errors; IPW is simpler to describe and does not need
an imputation model for the mediator, but depends on the missingness model and
discards incomplete rows. When they disagree by more than their standard errors,
suspect a misspecified model on one side.

## Limits

* `sensitivity_mnar()` is not available for `method = "ipw"`: delta adjustment
  shifts imputed values, and IPW has none.
* `infer(type = "mbco")` is not implemented for IPW. Use `type = "mc"`.
* `engine = "lavaan"` works with IPW, with robust standard errors only; see the
  *Structural equation models* article.
