---
title: "Inverse probability weighting, step by step"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Multiple imputation fills in the missing values. Inverse probability weighting
(IPW) does something different: it keeps only the **complete cases** and weights
each one by the inverse of its estimated probability of being observed, so the
complete cases stand in for everyone. missingmed runs both through the same
`set_md_mediation() -> run() -> pool() -> infer()` pipeline, selected by
`method`.

## Data with a missing mediator

Missingness in `M` depends on the covariate `C` (missing at random).

```{r data}
library(missingmed)
set.seed(1)
n <- 400
d <- data.frame(X = rbinom(n, 1, 0.5), C = rnorm(n))
d$M <- 0.5 * d$X + 0.3 * d$C + rnorm(n)
d$Y <- 0.3 * d$M + 0.2 * d$X + 0.3 * d$C + rnorm(n)
d$M[runif(n) < plogis(-1 + 0.8 * d$C)] <- NA
mean(is.na(d$M))
```

## Fit

For IPW, `data` is the raw data frame (not a `mids` object), and `method = "ipw"`.

```{r fit}
md <- set_md_mediation(d, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", method = "ipw"
)
fit <- run(md)
res <- pool(fit)
res
```

There is one fit (`m = 1`), so the pooled table is that fit's table, with robust
standard errors.

```{r infer}
set.seed(2)
infer(fit, type = "mc", n.mc = 20000)
```

## Look at the weights

`fit@weights` has one entry per row of the data: the weight for a complete case,
`NA` for an incomplete one.

```{r weights}
w <- fit@weights
c(complete = sum(!is.na(w)), incomplete = sum(is.na(w)))
summary(w)
```

A few very large weights dominate the fit. If the maximum is many times the
median, the missingness model has probabilities close to zero for some cases and
the estimate rests on them. Two remedies, both arguments of `set_md_mediation()`:
`weight_trim` caps the weights, and a simpler `weight_formula` avoids extreme
probabilities.

## Choosing the missingness model

By default the missingness model uses every observed predictor. `weight_formula`
overrides it: a single formula, or a named list with one formula per incomplete
variable.

```{r wf}
md2 <- set_md_mediation(d, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", method = "ipw",
  weight_formula = list(M = ~ X + C)
)
summary(run(md2)@weights)
```

## Stabilizing and trimming

`weight_stabilize = TRUE` (the default) multiplies by the marginal probability of
being observed, which leaves the estimate's target unchanged and shrinks the
weights' spread. `weight_trim` is an upper quantile: `0.95` caps the largest 5% of
weights at the 95th percentile. `1` (default) means no trimming.

```{r trim}
md3 <- set_md_mediation(d, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M", method = "ipw", weight_trim = 0.95
)
rbind(
  untrimmed = summary(run(md)@weights)[c("Min.", "Median", "Max.")],
  trimmed   = summary(run(md3)@weights)[c("Min.", "Median", "Max.")]
)
```

Trimming trades a little bias for less variance. Report whether you trimmed.

## Standard errors

`se_type = "sandwich"` (default) uses a heteroskedasticity-robust (HC) variance;
`"model"` uses the usual model-based one, which is wrong for weighted fits and is
there for comparison only. The weights are treated as **known**: the uncertainty
from estimating them is not propagated.

## IPW or imputation?

Compare them on the same data.

```{r compare}
imp <- mice::mice(d, m = 10, method = "norm", printFlag = FALSE, seed = 3)
mi <- pool(run(set_md_mediation(imp, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M"
)))
ipw <- res
cmp <- function(r) setNames(r@tidy_table$estimate, r@tidy_table$term)[c("a", "b", "c_prime")]
round(rbind(MI = cmp(mi), IPW = cmp(ipw)), 3)
```

Both rely on the missing-at-random assumption. Imputation uses every row and
usually has smaller standard errors; IPW is simpler to describe and does not need
an imputation model for the mediator, but depends on the missingness model and
discards incomplete rows. When they disagree by more than their standard errors,
suspect a misspecified model on one side.

## Limits

* `sensitivity_mnar()` is not available for `method = "ipw"`: delta adjustment
  shifts imputed values, and IPW has none.
* `infer(type = "mbco")` is not implemented for IPW. Use `type = "mc"`.
* `engine = "lavaan"` works with IPW, with robust standard errors only; see the
  *Structural equation models* article.
