---
title: "Structural equation models: the lavaan engine"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`set_md_mediation(engine = "lavaan")` runs a structural equation model on every
imputed dataset and pools the result with Rubin's rules. Use it when the
mediator is **latent** (measured by several indicators), when you want a
lavaan-specific option such as a robust estimator, or when the model is easier to
state as lavaan syntax than as two formulas. For one observed mediator and a
regression outcome the default `engine = "glm"` is simpler.

This tutorial builds up from an observed path model to a latent mediator.

## Data

An exposure `X`, a covariate `C`, a latent mediator measured by three noisy
indicators, and an outcome `Y`. Two of the indicators have missing values.

```{r data}
library(missingmed)
set.seed(51)
n <- 300
C <- rnorm(n)
X <- rbinom(n, 1, 0.5)
L <- 0.6 * X + 0.3 * C + rnorm(n)            # the latent mediator
d <- data.frame(
  X, C,
  m1 = L + rnorm(n, 0, 0.5),
  m2 = 0.8 * L + rnorm(n, 0, 0.5),
  m3 = 0.7 * L + rnorm(n, 0, 0.5)
)
d$Y <- 0.2 * X + 0.5 * L + 0.3 * C + rnorm(n)
d$m1[1:40] <- NA
d$m2[41:70] <- NA
imp <- suppressWarnings(
  mice::mice(d, m = 5, method = "norm", printFlag = FALSE, seed = 2)
)
```

## An observed path model

Give the lavaan syntax in `model`, and name the three roles. `outcome` is
required: it must be regressed on the mediator in `model`. (If two variables were
regressed on the mediator, medfit would silently take the first, so missingmed
asks you to say which.)

```{r observed}
d$Mobs <- (d$m1 + d$m2 + d$m3) / 3
imp_obs <- suppressWarnings(
  mice::mice(d, m = 5, method = "norm", printFlag = FALSE, seed = 2)
)
md_obs <- set_md_mediation(imp_obs,
  model = "Mobs ~ a * X + C\nY ~ b * Mobs + cp * X + C",
  treatment = "X", mediator = "Mobs", outcome = "Y",
  engine = "lavaan"
)
md_obs
```

The model is checked before anything is fitted: the syntax must parse, the three
roles must be distinct, `mediator ~ treatment` and `outcome ~ mediator` must be in
the model, and every observed variable must be a column of the data.

```{r observed-run}
res_obs <- pool(run(md_obs))
res_obs
```

The pooled `a`, `b` and `c_prime` equal what `engine = "glm"` gives on the same
imputations; see the *Reading the results* article for the columns.

## A latent mediator

Put the measurement model in the syntax and name the latent variable as the
`mediator`. It is not a data column, and that is fine.

```{r latent}
mod_lat <- "
  Mlat =~ m1 + m2 + m3
  Mlat ~ a * X + C
  Y    ~ b * Mlat + cp * X + C
"
md_lat <- set_md_mediation(imp,
  model = mod_lat,
  treatment = "X", mediator = "Mlat", outcome = "Y",
  engine = "lavaan"
)
fit_lat <- run(md_lat)
res_lat <- pool(fit_lat)
res_lat@tidy_table[, c("term", "estimate", "std_error", "statistic", "p_value")]
```

Two things to read in that table:

* The pooled **loadings** (`Mlat=~m2`, `Mlat=~m3`), with `m1` fixed to 1 by
  lavaan's default scaling, and the structural paths `a`, `b`, `cp`.
* The **variance rows** (`m1~~m1`, `Mlat~~Mlat`, ...) keep their estimate and
  standard error, but `statistic` and `p_value` are `NA`. A variance is tested
  against zero, which is on the boundary of its range, so a Wald z-test is not
  valid there. A covariance between two different variables (for example a
  residual covariance `Y~~Y2`) has an interior null, so it keeps its test.

The indirect effect and its Monte Carlo interval:

```{r latent-infer}
set.seed(1)
infer(fit_lat, type = "mc", n.mc = 20000)
```

## Passing options to lavaan

`fit_args` is a named list of extra arguments for `lavaan::sem()`. It is stored on
the object, so every later refit (including the ones inside
`sensitivity_mnar()`) uses it. `model` and `data` cannot be overridden.

```{r fit-args}
md_mlr <- set_md_mediation(imp,
  model = mod_lat, treatment = "X", mediator = "Mlat", outcome = "Y",
  engine = "lavaan", fit_args = list(estimator = "MLR")
)
md_mlr@fit_args
```

## What happens when a fit goes wrong

* **Non-convergence** in any imputation stops `run()` once, naming every
  imputation that failed. Simplify the model, or raise the iteration limit with
  `fit_args = list(control = list(iter.max = 5000))`.
* A **converged but improper solution** (for example a negative residual
  variance) is a warning, raised once and naming the imputations. Pooling goes on:
  inspect those imputations before trusting the result.

## Inverse probability weighting with lavaan

`method = "ipw"` also works with the lavaan engine. The complete cases are fitted
with `sampling.weights`, and the standard errors are always robust (sandwich), as
for the glm path. A request for model-based standard errors is refused rather than
silently replaced:

```{r ipw-refusal, error = TRUE}
set_md_mediation(d,
  model = "Mobs ~ a * X + C\nY ~ b * Mobs + cp * X + C",
  treatment = "X", mediator = "Mobs", outcome = "Y",
  engine = "lavaan", method = "ipw",
  fit_args = list(se = "standard")
)
```

## Sensitivity analysis with a latent mediator

`sensitivity_mnar()` shifts imputed values of a data column. A latent mediator has
no column, so name an observed indicator as `target`; the error lists the
indicators if you forget.

```{r sens-error, error = TRUE}
sensitivity_mnar(md_lat, delta = c(0, 0.5))
```

```{r sens}
s <- sensitivity_mnar(md_lat,
  delta = c(-0.5, 0, 0.5), target = "m1",
  n.mc = 2000, seed = 1
)
tidy(s)
```

## The MBCO test with a lavaan fit

`infer(type = "mbco")` runs the same D4-stacked likelihood-ratio test of
H0: a·b = 0 as the glm engine, with lavaan doing the refits. Each imputation is
fit once in full and once with the `a` path (`Mlat ~ X`) fixed to 0 and once with
the `b` path (`Y ~ Mlat`) fixed to 0; the stacked data are fit the same three
ways. With a latent mediator, only the **structural** path is fixed: the loadings,
the measurement errors and any direct effect of an indicator on `Y` stay free, so
each branch tests exactly one parameter (`k = 1`).

```{r mbco}
infer(fit_lat, type = "mbco")
```

Read it as for the glm engine (see `vignette("mbco-mi")`): `D4` and `p` are the
test, `r4` and `nu` describe the between-imputation variability, and
`branch_mix` reports whether the imputations disagreed about which path is the
weaker one. Two properties to know:

* **It is conservative at a = b = 0.** The statistic is the smaller of two
  likelihood-ratio statistics, so when both paths are null the test rejects less
  often than its nominal level. That costs power near the intersection; it does
  not inflate false positives.
* **It does not test direct effects of the indicators.** Those are not part of
  `a * b`, so a nonzero direct effect of an indicator on `Y` is left in the model
  under both nulls.

The same test is available for a plain list of completed data frames, without a
`set_md_mediation()` object:

```{r mbco-d4, eval = FALSE}
mbco_d4(mice::complete(imp, "all"),
  model = mod_lat, treatment = "X", mediator = "Mlat", outcome = "Y"
)
```

`sensitivity_mnar(type = "mbco")` works too, and returns one MBCO result per
rung.

### What it does not support

MBCO is refused, naming the option, for `estimator` other than `"ML"` (`"MLR"`,
`"MLM"`, `"WLSMV"`, ...), for `group`, `ordered` and `sampling.weights`, and for
`method = "ipw"`. Robust (scaled) test statistics are excluded because the D4
pooling is justified for likelihood-ratio statistics, not for scaled ones, and no
validated pooling exists yet. Use `type = "mc"` for those fits.

A refit that does not converge stops the test, naming the imputation (or the
stacked data) and which fit; an improper solution warns once and goes on.
