---
title: "Reading the pooled results"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`pool()` returns an `MDMediationResult`. This article reads its table column by
column, shows what `infer()` adds, and says what each number does and does not
tell you.

```{r data}
library(missingmed)
set.seed(1)
n <- 300
d <- data.frame(X = rbinom(n, 1, 0.5), C = rnorm(n))
d$M <- 0.5 * d$X + 0.3 * d$C + rnorm(n)
d$Y <- 0.3 * d$M + 0.2 * d$X + 0.3 * d$C + rnorm(n)
d$M[runif(n) < plogis(-1 + 0.8 * d$C)] <- NA
imp <- mice::mice(d, m = 10, method = "norm", printFlag = FALSE, seed = 1)
md <- set_md_mediation(imp, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M",
  conf_int = TRUE, conf_level = 0.95
)
fit <- run(md)
res <- pool(fit)
```

## The table

```{r table}
tab <- res@tidy_table
names(tab)
```

| Column | Meaning |
|---|---|
| `term` | Coefficient. The mediator model's terms start `m_`, the outcome model's `y_`; `a`, `b`, `c_prime` are the three path aliases. |
| `estimate` | Rubin's pooled point estimate: the mean over imputations. |
| `std_error` | Square root of the total variance. |
| `var_w`, `var_b`, `var_tot` | Within-imputation, between-imputation and total variance, with `var_tot = var_w + (1 + 1/m) var_b`. |
| `statistic` | `estimate / std_error`. |
| `df` | Degrees of freedom for the reference distribution (below). |
| `riv` | Relative increase in variance due to missing data. |
| `fmi` | Fraction of missing information. |
| `p_value` | Two-sided p-value on that reference distribution. |
| `conf_low`, `conf_high` | Only if `conf_int = TRUE`, at `conf_level`. |

```{r paths}
tab[tab$term %in% c("a", "b", "c_prime"),
  c("term", "estimate", "std_error", "df", "riv", "fmi", "p_value", "conf_low", "conf_high")]
```

## Degrees of freedom

With more than one imputation the reference is Student's *t* with the
Barnard–Rubin degrees of freedom. It uses the complete-data degrees of freedom of
the model the coefficient belongs to (`n_obs` minus the number of that model's
terms), so a small sample is not treated as an infinite one. For binomial and
Poisson models, and for lavaan fits, the complete-data degrees of freedom are
infinite and the reference is a large-sample *t* (with `m > 1`) or a normal
(with `m = 1`).

With a single fit (`m = 1`) there is no between-imputation variance: `riv` and
`fmi` are zero and `df` is the complete-data value, so the table is the ordinary
Wald table of that fit.

## `riv` and `fmi`

`riv` is how much larger the variance is because of the missing data; `fmi` is the
share of the total information that is missing. Large values (say `fmi` above
0.5) mean the answer leans heavily on the imputation model. They are estimated
from `m` imputations and are noisy for small `m`: with `m = 5`, treat them as
rough.

```{r fmi}
round(range(tab$fmi), 3)
```

## Confidence intervals of single coefficients

`conf_low` and `conf_high` bound **one coefficient** on Rubin's *t* reference.
They are not an interval for the indirect effect, which is a product of two
coefficients and is not normally distributed.

## The indirect effect: `infer()`

```{r infer}
set.seed(2)
mc <- infer(res, type = "mc", n.mc = 20000)
mc
```

`type = "mc"` draws from the pooled distribution of `(a, b)` and reports the
percentile interval of `a * b` (via RMediation). `$CI` is the interval,
`$Estimate` the point estimate, `$SE` the Monte Carlo standard error of the
product, and `$MC.Error` the simulation error: raise `n.mc` to shrink it.

`level` defaults to the `conf_level` you set on the data object; an explicit
`level =` in `infer()` wins.

## A test instead of an interval

The D4-stacked MBCO test needs the per-imputation fits, so call it on the fit, not
on the pooled result:

```{r mbco}
mb <- infer(fit, type = "mbco")
mb
```

`mb["p"]` is the p-value, `mb["D4"]` the statistic, `mb["nu"]` its
denominator degrees of freedom. See the *Testing an indirect effect* article.

## What is not in the table

* A p-value for the **indirect effect** itself: use `infer()`.
* A test of the missing-at-random assumption: none exists, and missingmed does not
  pretend to give one. `sensitivity_mnar()` shows how far a conclusion moves when
  MAR fails.
* For lavaan fits, the variance and covariance rows (`~~`) have `NA` statistic and
  p-value by design; see the *Structural equation models* article.
