---
title: "Migrating from the S4 API"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The S4 functions `set_sem()`, `run_sem()` and `pool_sem()` and their classes
are deprecated. They still run, with a deprecation warning, and they are planned
to become `.Defunct()` stubs in 0.6.0, each naming its replacement. This
article maps the old calls to the S7 pipeline.

## The mapping

| S4 (deprecated) | S7 | Returns |
|---|---|---|
| `set_sem(data, model)` | `set_md_mediation(data, formula_y, formula_m, treatment, mediator)` | `MDMediationData` |
| `run_sem(object)` | `run(object)` | `MDMediationFit` |
| `pool_sem(object)` | `pool(object)` | `MDMediationResult` |
| (none) | `infer(fit, type = "mc" or "mbco")` | interval or test for the indirect effect |
| `SemImputedData`, `SemResults`, `PooledSEMResults` | `MDMediationData`, `MDMediationFit`, `MDMediationResult` | |

## From model syntax to two formulas

The S4 API took a lavaan model string or an OpenMx model. The S7 API takes the
mediator model and the outcome model as two formulas, plus the names of the
treatment and the mediator. An observed-variable lavaan model

```{r, eval = FALSE}
model <- "
  M ~ a * X + C
  Y ~ b * M + cp * X + C
"
res_s4 <- pool_sem(run_sem(set_sem(imp, model)))
```

becomes

```{r}
library(missingmed)
set.seed(4)
n <- 300
d <- data.frame(X = rbinom(n, 1, 0.5), C = rnorm(n))
d$M <- 0.5 * d$X + 0.3 * d$C + rnorm(n)
d$Y <- 0.3 * d$M + 0.2 * d$X + 0.3 * d$C + rnorm(n)
d$M[sample(n, 40)] <- NA
imp <- mice::mice(d, m = 5, method = "norm", printFlag = FALSE, seed = 4)

md <- set_md_mediation(imp,
  formula_y = Y ~ X + M + C,
  formula_m = M ~ X + C,
  treatment = "X", mediator = "M"
)
fit <- run(md)
res <- pool(fit)
res
```

The path labels are fixed rather than user-chosen: `a`, `b` and `c_prime` in
the pooled estimates, plus `theta3` and `b0` for an `X:M` model. The pooled
table is `res@tidy_table`.

## What the S7 pipeline adds

* `infer()`: a Monte Carlo interval for the indirect effect, or the D4-MBCO
  test, where the S4 API pooled path coefficients only.
* `method = "ipw"` for inverse probability weighting.
* `sensitivity_mnar()` for departures from MAR.
* Validation of the model before fitting (see the *Supported models* article).

## What has no replacement yet

* **Latent mediators and other SEM features.** A lavaan engine for
  `set_md_mediation()` is planned for 0.6.0.
* **OpenMx models.** No OpenMx engine is planned for 0.6.0.

Until then, these models still run through the deprecated S4 functions.

## Silencing the warning while you migrate

The deprecation warning is an ordinary warning, so a script that must keep the
S4 call for now can wrap it in `suppressWarnings()`. Note that `pool_sem()`
p-values use Rubin's t reference with Barnard–Rubin df since 0.5.1; results
from earlier versions can differ.
