---
title: "Sensitivity analysis for departures from MAR"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Every analysis in this package assumes the data are **missing at random** (MAR):
given what you observed, missingness does not depend on the value that is
missing. That assumption cannot be tested from the data. `sensitivity_mnar()`
asks a different question: *if the missing values were systematically higher or
lower than MAR implies, how much would the conclusion change?* It produces a
sensitivity curve, not an estimate. Nothing it reports is identified.

```{r data}
library(missingmed)
set.seed(1)
n <- 300
d <- data.frame(X = rbinom(n, 1, 0.5), C = rnorm(n))
d$M <- 0.5 * d$X + 0.3 * d$C + rnorm(n)
d$Y <- 0.3 * d$M + 0.2 * d$X + 0.3 * d$C + rnorm(n)
d$M[runif(n) < plogis(-1 + 0.8 * d$C)] <- NA
imp <- mice::mice(d, m = 5, method = "norm", printFlag = FALSE, seed = 1)
md <- set_md_mediation(imp, Y ~ X + M + C, M ~ X + C,
  treatment = "X", mediator = "M"
)
```

## One shift

A shift `delta` is added to the imputed values of the `target` (the mediator by
default), in the variable's own units, and the analysis is repeated. A grid of
shifts gives a curve; `delta = 0` is the MAR analysis.

```{r curve}
s <- sensitivity_mnar(md, delta = c(-0.5, 0, 0.5), n.mc = 2000, seed = 1)
tidy(s)
```

Read `estimate`, `conf_low` and `conf_high` down the rungs: the indirect effect
and its interval at each delta. If the interval keeps the same sign across a
plausible range of deltas, the conclusion is robust to that departure.

## `delta` is conditional; `msp` is what happened

`delta` is a **conditional** shift (given the observed predictors). The column
`msp` is the *marginal* difference actually realized between imputed and observed
values. Compare `msp` with what you meant: a conditional shift of 0.5 does not
make the marginal shift 0.5.

```{r msp}
tidy(s)[, c("M", "msp")]
```

## Choosing the range

There is no data-driven range. Choose deltas from substantive knowledge: how
different could the missing mediator values plausibly be from the observed ones,
in units of the mediator? Then report the whole curve. `summary(s)` also finds a
**tipping point** for a numeric grid: the smallest shift at which the conclusion
changes.

```{r summary}
summary(s)
```

Here no shift in the grid changes the conclusion, which is the reassuring
outcome: the interval keeps its sign across the whole range tried.

## Several variables at once

Pass a data frame of shifts, one column per variable. Leave `target` as `NULL`;
the column names are the targets.

```{r multi}
sensitivity_mnar(md, delta = data.frame(M = c(0, 0.5)), n.mc = 2000, seed = 1)
```

(One column here because only `M` has missing values in this example; with
several incomplete variables, add one column per variable.)

## A shift that depends on a covariate

`ums` replaces `delta` with a covariate-varying offset, passed to `mice`'s NARFCS
imputation methods. Each string is one rung and needs exactly one intercept term,
for example `"0.5 + 0.2*C"`: the offset is `0.5 + 0.2 C` for each row. A `ums`
grid has no numeric order, so `summary()` computes no tipping point for it.

```{r ums, eval = FALSE}
sensitivity_mnar(md, ums = c("0", "0.5 + 0.2*C"), n.mc = 2000)
```

## A test per rung

`type = "mbco"` runs the D4 test at each delta instead of the interval. The
tipping point is then the delta at which the test stops rejecting.

```{r mbco, eval = FALSE}
sensitivity_mnar(md, delta = c(-0.5, 0, 0.5), type = "mbco")
```

## What it assumes, and what it does not cover

* The imputation model you supplied is compatible with the mediation model. That
  is not verifiable.
* The seed is pinned across rungs (`seed`, defaulting to the one stored in the
  `mids` object), so differences between rungs are the delta, not Monte Carlo
  noise.
* `method = "mi"` only. IPW has no imputed values to shift.
* With `engine = "lavaan"`, `type = "mc"` works. A latent mediator needs an
  explicit observed `target`; see the *Structural equation models* article.
