---
title: "Supported models"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`set_md_mediation()` checks the model before anything is fitted. This article
lists what it accepts and what it refuses, and runs every rule below: each
refusal is asserted, so the page fails to build if a rule changes.

```{r setup}
library(missingmed)
set.seed(21)
n <- 300
d <- data.frame(X = rbinom(n, 1, 0.5), C = rnorm(n), W = rnorm(n))
d$M <- 0.5 * d$X + 0.3 * d$C + rnorm(n)
d$Y <- 0.3 * d$M + 0.2 * d$X + 0.3 * d$C + rnorm(n)
d$M[sample(n, 40)] <- NA
imp <- mice::mice(d, m = 3, method = "norm", printFlag = FALSE, seed = 21)

# TRUE when set_md_mediation() accepts the specification; otherwise prints the
# refusal and returns FALSE.
accepts <- function(...) {
  tryCatch({
    set_md_mediation(...)
    TRUE
  }, error = function(e) {
    cat("Refused:", conditionMessage(e), "\n")
    FALSE
  })
}
```

## The model

Two regressions with named roles:

* the **mediator model** `formula_m`, whose response is the mediator and which
  contains the treatment as a main effect;
* the **outcome model** `formula_y`, which contains the treatment and the
  mediator as main effects.

Covariates enter either model freely. Formulas are expanded against the data
first, so `Y ~ .` is checked as the model that `run()` fits.

## Accepted

```{r accepted}
stopifnot(
  # the basic model
  accepts(imp, Y ~ X + M + C, M ~ X + C, treatment = "X", mediator = "M"),
  # a treatment-by-mediator interaction, in any spelling
  accepts(imp, Y ~ X * M + C, M ~ X + C, treatment = "X", mediator = "M"),
  accepts(imp, Y ~ M:X + X + M + C, M ~ X + C, treatment = "X", mediator = "M"),
  # covariate transforms and covariate products
  accepts(imp, Y ~ X + M + poly(C, 2) + C:W, M ~ X + C, treatment = "X", mediator = "M"),
  # a binary outcome or mediator (no X:M term)
  accepts(imp, I(Y > 0) ~ X + M + C, M ~ X + C, treatment = "X", mediator = "M",
    family_y = binomial())
)
```

The treatment may be binary or continuous, as long as the column is numeric.
The outcome's response may be a transform (`I(Y > 0)` above); the mediator's
may not, as shown next.

## Refused

Each rule below names the term or argument at fault.

```{r refused}
stopifnot(
  # formula_m must model the mediator itself
  !accepts(imp, Y ~ X + M + C, C ~ X, treatment = "X", mediator = "M"),
  # a transformed mediator needs its own column
  !accepts(imp, Y ~ X + M + C, log(M + 10) ~ X + C, treatment = "X", mediator = "M"),
  # the treatment must be a main effect of the outcome model
  !accepts(imp, Y ~ M + C, M ~ X + C, treatment = "X", mediator = "M"),
  # products other than X:M
  !accepts(imp, Y ~ X + M + X:C, M ~ X + C, treatment = "X", mediator = "M"),
  !accepts(imp, Y ~ X + M * W, M ~ X + C, treatment = "X", mediator = "M"),
  # transforms of the treatment or mediator
  !accepts(imp, Y ~ X + M + I(X^2), M ~ X + C, treatment = "X", mediator = "M"),
  # X:M outside a Gaussian identity-link model
  !accepts(imp, I(Y > 0) ~ X * M + C, M ~ X + C, treatment = "X", mediator = "M",
    family_y = binomial())
)
```

A treatment stored as a factor, character or logical column is refused too;
recode it to numeric (0/1 for a binary treatment):

```{r refused-factor}
d_f <- d
d_f$X <- factor(d_f$X, labels = c("control", "treated"))
meth <- mice::make.method(d_f)
meth["M"] <- "norm"
imp_f <- mice::mice(d_f, m = 2, method = meth, printFlag = FALSE, seed = 1)
stopifnot(!accepts(imp_f, Y ~ X + M + C, M ~ X + C, treatment = "X", mediator = "M"))
```

## Summary of the rules

| Rule | Example refused |
|---|---|
| `formula_m`'s response is the bare mediator column | `C ~ X`, `log(M) ~ X` |
| `formula_y`'s response does not involve the mediator | `M ~ X + C` as the outcome model |
| the treatment is a main effect of both models | `Y ~ M + C` |
| the mediator is a main effect of the outcome model | `Y ~ X + C` |
| treatment and mediator appear only as main effects, plus one `X:M` in the outcome model | `X:C`, `M:W`, `X:M:W`, `I(X^2)`, `poly(X, 2)` |
| `X:M` only with Gaussian identity-link models | `X:M` with `family_y = binomial()` |
| the treatment column is numeric | factor, character, logical |
| `treatment` and `mediator` are syntactic R names | `"my M"` |
| `engine` is `"glm"`, or `"regmedint"` with medfit >= 0.4.0 and `method = "mi"` | `engine = "lavaan"` |

`?set_md_mediation` lists the same rules under *Details*, together with the
checks on IPW's `weight_formula`.

## Moderated models

Models that `set_md_mediation()` refuses because of a product term can still be
tested for mediation with `mbco_d4()`, which runs D4-MBCO on a plain list of
completed datasets and accepts moderated models such as `Y ~ X + M * W`. See
`vignette("mbco-mi")`.

## Latent mediators

Structural equation models are not in the S7 pipeline yet. The deprecated S4
API (`set_sem()`) still fits lavaan and OpenMx models; a lavaan engine for
`set_md_mediation()` is planned for 0.6.0. See the *Migrating from the S4 API*
article.
